---
title: "Gut microbiome composition and economic preferences"
subtitle: "Reproducible analysis notebook (public-repository version; based on v5.3c)"
author: "Behavioural microbiome manuscript team"
date: "`r Sys.Date()`"
params:
  data_dir: "data"
  output_dir: "outputs"
  run_id: ""
  run_maaslin: true
  run_shotgun_species_maaslin: true
  run_fb_ratio: true
  run_metabolomics: true
  maaslin_cores: 1
  permutations: 9999
  random_seed: 20260721
  source_run_dir: ""
  top_nominal_per_panel: 12
  strict_checks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
    df_print: paged
---

# Purpose

This stable v5.3c notebook is the single analysis source for the microbiome--economic-preferences manuscript. It consolidates the last working data-loading, alpha-diversity, PERMANOVA, MaAsLin2, taxon-figure, and metabolomics code while retaining the corrections agreed during manuscript review.

It is designed to be run section by section in RStudio Server. Each section should:

1. read or compute the relevant result,
2. write reproducible output tables,
3. write publication-ready figure files,
4. print a short manuscript-ready interpretation.

The notebook uses the cohort labels **Community cohort** and **Student cohort** throughout. Internal file names may still contain `hamburg` or `american` for compatibility with existing outputs. Species-level MaAsLin2 is restricted to the Community-cohort shotgun data; the Student-cohort 16S data are analysed at family and genus level only. The harmonized Bacillota/Bacteroidota ratio is reinstated as a secondary exploratory analysis using cohort-specific rarefied phylum tables and is retained whether its result is positive or null.

For public-repository use, input data are intentionally separated from code. The repository does not contain participant-level or otherwise restricted source data. Input locations are configured via `data_dir`/`MICROBIOME_PREF_DATA`, and generated outputs are written below `outputs/`, which is excluded from version control by default.

# 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300,
  cache = FALSE
)

# R Markdown parameters are automatically available when knitting, but not
# always when “Run All” is used interactively in RStudio Server.
get_notebook_param <- function(name, default) {
  if (exists("params", inherits = TRUE)) {
    p <- get("params", inherits = TRUE)
    if (!is.null(p[[name]])) return(p[[name]])
  }
  default
}

# Public-repository path handling. No institution- or user-specific paths are
# embedded in the analysis. By default, inputs are expected under data/ and
# outputs are written under outputs/ relative to the repository root.
# Restricted data may instead be stored anywhere and supplied via
# MICROBIOME_PREF_DATA=/path/to/private/data.
PROJECT_DIR <- normalizePath(
  Sys.getenv("MICROBIOME_PREF_PROJECT", unset = getwd()),
  mustWork = FALSE
)

resolve_project_path <- function(path, base = PROJECT_DIR) {
  if (is.null(path) || !nzchar(path)) return(NULL)
  is_abs <- startsWith(path, "/") || grepl("^[A-Za-z]:", path)
  normalizePath(if (is_abs) path else file.path(base, path), mustWork = FALSE)
}

data_param <- get_notebook_param("data_dir", "data")
output_param <- get_notebook_param("output_dir", "outputs")

DATA_DIR <- resolve_project_path(
  Sys.getenv("MICROBIOME_PREF_DATA", unset = data_param)
)
OUTPUT_DIR <- resolve_project_path(
  Sys.getenv("MICROBIOME_PREF_OUTPUT", unset = output_param)
)

RUN_ID <- get_notebook_param("run_id", "")
if (is.null(RUN_ID) || !nzchar(RUN_ID)) RUN_ID <- format(Sys.time(), "%Y%m%d_%H%M%S")
RUN_MAASLIN <- isTRUE(get_notebook_param("run_maaslin", TRUE))
RUN_SPECIES_MAASLIN <- isTRUE(get_notebook_param(
  "run_shotgun_species_maaslin",
  get_notebook_param("run_species_maaslin", TRUE)
))
RUN_FB_RATIO <- isTRUE(get_notebook_param("run_fb_ratio", TRUE))
RUN_METABOLOMICS <- isTRUE(get_notebook_param("run_metabolomics", TRUE))
MAASLIN_CORES <- as.integer(get_notebook_param("maaslin_cores", 1L))
PERMUTATIONS <- as.integer(get_notebook_param("permutations", 9999L))
RANDOM_SEED <- as.integer(get_notebook_param("random_seed", 20260721L))
set.seed(RANDOM_SEED)
SOURCE_RUN_DIR <- get_notebook_param("source_run_dir", "")
if (is.null(SOURCE_RUN_DIR) || !nzchar(SOURCE_RUN_DIR)) SOURCE_RUN_DIR <- NULL
if (!is.null(SOURCE_RUN_DIR)) SOURCE_RUN_DIR <- resolve_project_path(SOURCE_RUN_DIR)
TOP_NOMINAL_PER_PANEL <- as.integer(get_notebook_param("top_nominal_per_panel", 12L))
STRICT_CHECKS <- isTRUE(get_notebook_param("strict_checks", TRUE))
NOTEBOOK_VERSION <- "v5_3c_student_stage2_covariates_2026-09-23"
PDF_DEVICE <- if (capabilities("cairo")) grDevices::cairo_pdf else grDevices::pdf
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(readxl)
  library(vegan)
  library(broom)
  library(janitor)
  library(stringr)
  library(forcats)
  library(patchwork)
})

# MaAsLin2 is only loaded if needed because it can be slow / unavailable on some systems.
if (RUN_MAASLIN) {
  suppressPackageStartupMessages(library(Maaslin2))
}
```

```{r directories}
if (!dir.exists(DATA_DIR)) {
  stop(
    "DATA_DIR does not exist: ", DATA_DIR, "\n",
    "Create data/ in the repository or set MICROBIOME_PREF_DATA to the private input-data directory.",
    call. = FALSE
  )
}
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
RUN_DIR <- file.path(OUTPUT_DIR, RUN_ID)
RESULTS_DIR <- file.path(RUN_DIR, "results")
TABLES_DIR  <- file.path(RUN_DIR, "tables")
FIGURES_DIR <- file.path(RUN_DIR, "figures")
SUPP_DIR    <- file.path(RUN_DIR, "supplement")
LOG_DIR     <- file.path(RUN_DIR, "logs")

purrr::walk(
  c(RUN_DIR, RESULTS_DIR, TABLES_DIR, FIGURES_DIR, SUPP_DIR, LOG_DIR),
  ~ dir.create(.x, recursive = TRUE, showWarnings = FALSE)
)

cat("PROJECT_DIR:", PROJECT_DIR, "\n")
cat("DATA_DIR:", DATA_DIR, "\n")
cat("OUTPUT_DIR:", OUTPUT_DIR, "\n")
cat("RUN_DIR:", RUN_DIR, "\n")
cat("NOTEBOOK_VERSION:", NOTEBOOK_VERSION, "\n")
cat("RUN_MAASLIN:", RUN_MAASLIN, "\n")
cat("RUN_SPECIES_MAASLIN (Community shotgun only):", RUN_SPECIES_MAASLIN, "\n")
cat("RUN_FB_RATIO:", RUN_FB_RATIO, "\n")
cat("RUN_METABOLOMICS:", RUN_METABOLOMICS, "\n")
cat("RANDOM_SEED:", RANDOM_SEED, "\n")
cat("PDF_DEVICE:", if (capabilities("cairo")) "cairo_pdf" else "pdf", "\n")
writeLines(NOTEBOOK_VERSION, file.path(LOG_DIR, "notebook_version.txt"))
```

# 1. Global configuration and helper functions

```{r configuration}
COHORT_LABELS <- c(
  hamburg = "Community cohort",
  american = "Student cohort",
  Hamburg = "Community cohort",
  American = "Student cohort"
)

pretty_cohort <- function(x) {
  out <- unname(COHORT_LABELS[match(x, names(COHORT_LABELS))])
  ifelse(is.na(out), x, out)
}

PREFERENCE_CANDIDATES <- c(
  "donation", "risktaking", "timediscounting", "trust", "selfcontrol",
  "religiousness", "conservative", "liberal", "social", "ecological",
  "negative_reciprocity", "positive_reciprocity"
)

# Central labels and colour scales. Every exported figure uses these objects.
COHORT_COLORS <- c(
  "Community cohort" = "#F8766D",
  "Student cohort" = "#00BFC4"
)

PREFERENCE_LABELS <- c(
  "z_donation" = "Donation",
  "z_risktaking" = "Risk-taking",
  "z_timediscounting" = "Patience",
  "z_trust" = "Trust",
  "z_selfcontrol" = "Self-control",
  "z_religiousness" = "Religiousness",
  "z_conservative" = "Conservative orientation",
  "z_liberal" = "Liberal orientation",
  "z_social" = "Social orientation",
  "z_ecological" = "Ecological orientation",
  "z_negative_reciprocity" = "Negative reciprocity",
  "z_positive_reciprocity" = "Positive reciprocity"
)

PREFERENCE_COLORS <- c(
  "Donation" = "#D89000",
  "Risk-taking" = "#E76F51",
  "Patience" = "#0072B2",
  "Trust" = "#6A5ACD",
  "Self-control" = "#E83E8C",
  "Religiousness" = "#8C564B",
  "Conservative orientation" = "#F8766D",
  "Liberal orientation" = "#00A676",
  "Social orientation" = "#56B4E9",
  "Ecological orientation" = "#7CAE00",
  "Negative reciprocity" = "#00BFC4",
  "Positive reciprocity" = "#B983FF"
)

COVARIATE_COLORS <- c(
  "Age" = "#4D4D4D",
  "Gender" = "#8C8C8C",
  "BMI" = "#1F1F1F",
  "Smoking" = "#8C6D31",
  "Alcohol" = "#A6761D",
  "Alcohol frequency" = "#A6761D",
  "Club membership" = "#B3B3B3",
  "Other covariate" = "#B3B3B3"
)

ASSOCIATION_COLORS <- c(PREFERENCE_COLORS, COVARIATE_COLORS)

ECOLOGICAL_QUARTILE_COLORS <- c(
  "Q1 lowest" = "#F8766D",
  "Q2" = "#7CAE00",
  "Q3" = "#00BFC4",
  "Q4 highest" = "#C77CFF",
  "Missing" = "#8C8C8C"
)

# The linked metadata use code 1 for female participants in both cohorts.
# Keeping the codes explicit makes the cohort-description table auditable.
FEMALE_CODE <- c(community = "1", student = "1")

# Questionnaire items were administered on 0--10 scales, but the stored
# behavioural variables are coded 1--11. This offset is used ONLY for
# descriptive reporting (e.g. Table 1). Inferential models continue to use
# within-cohort z-scores, for which subtracting a constant has no effect.
# Donation is a 0--50 EUR amount and is never shifted.
PREFERENCE_DESCRIPTIVE_OFFSET <- c(community = 1, student = 1)

METABOLITE_COLORS <- c(
  "3-IAA" = unname(PREFERENCE_COLORS["Self-control"]),
  "Other metabolites" = "#8C8C8C"
)

FDR_THRESHOLD <- 0.10
NOMINAL_THRESHOLD <- 0.05

CORE_COVARIATES <- c("age", "gender", "bmi")

LIFESTYLE_CANDIDATES_COMMUNITY <- c(
  "smoking", "alcohol", "dietary_fibre_g", "saturated_fa_g",
  "income", "clubmember", "marital_status"
)

LIFESTYLE_CANDIDATES_STUDENT <- c(
  "smoking", "clubmember", "alcohol_frequency"
)

ALIASES <- list(
  bmi = c("bmi", "body_mass_index", "bodymassindex"),
  age = c("age", "age_years"),
  gender = c("gender", "sex"),
  marital_status = c("marital_status", "martial_status", "maritalstatus", "married"),
  smoking = c("smoking", "smoke", "smoking_status", "smoking_frequency"),
  alcohol = c("alcohol", "alcohol_consumption", "drinking"),
  alcohol_frequency = c("alcohol_frequency", "alcohol_freq", "drinking_frequency"),
  clubmember = c("clubmember", "club_member", "club_membership"),
  donation = c("donation", "donate", "giving"),
  risktaking = c("risktaking", "risk_taking", "risk"),
  timediscounting = c("timediscounting", "time_discounting", "discounting", "temporal_discounting"),
  selfcontrol = c("selfcontrol", "self_control"),
  trust = c("trust"),
  religiousness = c("religiousness", "religiosity"),
  conservative = c("conservative"),
  liberal = c("liberal"),
  social = c("social"),
  ecological = c("ecological"),
  negative_reciprocity = c("negative_reciprocity"),
  positive_reciprocity = c("positive_reciprocity")
)

TAXON_SYNONYMS <- c(
  Firmicutes = "Bacillota",
  Bacteroidetes = "Bacteroidota",
  Actinobacteria = "Actinomycetota",
  Proteobacteria = "Pseudomonadota"
)

cohort_config <- list(
  community = list(
    cohort = "Community cohort",
    internal = "hamburg",
    metadata_file = "23622_meta_dummy.r",
    preference_file = "Hamburg_AltruismVariables.csv",
    feature_tables = list(
      family = "HHaf_motusv3.1_family.tsv",
      genus = "HHaf_motusv3.1_genus.tsv",
      species = "HHaf_motusv3.1.tsv"
    ),
    rarefied_tables = list(
      family = "rtk_lowrc_HHaf_motusv3.1_family.tsvrarefied_to_8902.000000_n_0.tsv",
      genus = "rtk_lowrc_HHaf_motusv3.1_genus.tsvrarefied_to_8902.000000_n_0.tsv",
      phylum = "rtk_lowrc_HHaf_motusv3.1_phylum.tsvrarefied_to_8902.000000_n_0.tsv",
      fine = "rtk_lowrc_HHaf_motusv3.1.tsvrarefied_to_8902.000000_n_0.tsv"
    ),
    # One alpha-diversity table per cohort is used for manuscript/QC plots.
    # This avoids accidentally mixing RTK outputs from different taxonomic levels.
    alpha_table = "rtk_lowrc_HHaf_motusv3.1.tsvmedian_alpha_diversity.tsv",
    stage1 = CORE_COVARIATES,
    stage2 = LIFESTYLE_CANDIDATES_COMMUNITY,
    metabolomics = TRUE,
    metabolome_file = "metabolome.r"
  ),
  student = list(
    cohort = "Student cohort",
    internal = "american",
    metadata_file = "UHHdata.xlsx",
    preference_file = NULL,
    feature_tables = list(
      family = "Family.txt",
      genus = "Genus.txt"
    ),
    rarefied_tables = list(
      family = "rtkFamilyamgrarefied_to_7572.000000_n_0.tsv",
      genus = "rtkGenusamgrarefied_to_7572.000000_n_0.tsv",
      phylum = "lowrc_Phylumrtkrarefied_to_7423.000000_n_0.tsv"
    ),
    # Student cohort alpha table. Do NOT use OTUrtkmedian_alpha_diversity.tsv here;
    # that file has implausibly low richness values and is kept only for diagnostics.
    alpha_table = "rtkmedian_alpha_diversity.tsv",
    stage1 = CORE_COVARIATES,
    stage2 = LIFESTYLE_CANDIDATES_STUDENT,
    metabolomics = FALSE,
    metabolome_file = NULL
  )
)
```

```{r helper-functions}
alias_map <- setNames(rep(names(ALIASES), lengths(ALIASES)), unlist(ALIASES, use.names = FALSE))

canon <- function(x) {
  x <- stringr::str_trim(tolower(as.character(x)))
  x <- stringr::str_replace_all(x, "[^a-z0-9]+", "_")
  x <- stringr::str_replace_all(x, "_+", "_")
  x <- stringr::str_replace_all(x, "^_|_$", "")
  x <- ifelse(x %in% names(alias_map), alias_map[x], x)
  make.unique(x, sep = "_")
}

read_any <- function(path, sheet = 1) {
  if (is.null(path) || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) {
    return(as.data.frame(
      readr::read_tsv(path, show_col_types = FALSE, progress = FALSE, name_repair = "minimal"),
      check.names = FALSE
    ))
  }
  if (ext == "csv") {
    return(as.data.frame(
      readr::read_csv(path, show_col_types = FALSE, progress = FALSE, name_repair = "minimal"),
      check.names = FALSE
    ))
  }
  if (ext %in% c("xlsx", "xls")) {
    return(as.data.frame(
      readxl::read_excel(path, sheet = sheet, .name_repair = "minimal"),
      check.names = FALSE
    ))
  }
  stop("Unsupported file type: ", path)
}

coerce_id <- function(x) {
  x <- as.character(x)
  x <- stringr::str_trim(x)
  x <- basename(x)
  x <- stringr::str_replace_all(x, "\\.0+$", "")
  x <- stringr::str_replace_all(x, "^X+", "")

  # Prefer a terminal D-number if present.
  d_id <- stringr::str_extract(x, "D\\d+$")

  # Otherwise use the last numeric sample-like suffix.
  # IMPORTANT: allow one-digit IDs. The previous {2,8} pattern left
  # preference IDs 1, 3, 5, 7 and 9 as bare numbers instead of mapping
  # them to D1, D3, D5, D7 and D9, producing five false non-matches.
  n_id <- stringr::str_extract(x, "\\d{1,8}$")
  n_id <- stringr::str_replace_all(n_id, "^0+", "")

  out <- dplyr::case_when(
    !is.na(d_id) ~ d_id,
    !is.na(n_id) & n_id != "" ~ paste0("D", n_id),
    TRUE ~ x
  )

  out
}

numify <- function(x) {
  if (is.numeric(x) || is.integer(x)) return(as.numeric(x))
  suppressWarnings(readr::parse_number(as.character(x)))
}

zscore <- function(x) {
  x <- numify(x)
  if (all(is.na(x))) return(x)
  sx <- stats::sd(x, na.rm = TRUE)
  if (!is.finite(sx) || sx == 0) return(rep(0, length(x)))
  as.numeric(scale(x))
}

find_col <- function(df, candidates) {
  n <- canon(names(df))
  hit <- which(n %in% canon(candidates))
  if (length(hit)) names(df)[hit[1]] else NULL
}

find_best_col <- function(df, candidates, pattern = NULL) {
  # Prefer the candidate column with the most non-missing information.
  # This is important for smoking in the Student cohort: depending on prior
  # name harmonization, the informative column can be called smoking,
  # smoking_frequency, smoke, smoking_status, or smoking_1.
  cn <- names(df)
  can <- canon(cn)
  candidate_can <- canon(candidates)

  idx <- which(can %in% candidate_can)

  if (!is.null(pattern)) {
    idx <- union(idx, grep(pattern, cn, ignore.case = TRUE))
    idx <- union(idx, grep(pattern, can, ignore.case = TRUE))
  }

  idx <- idx[!is.na(idx) & idx >= 1 & idx <= length(cn)]
  idx <- unique(idx)
  # Never select audit/helper fields such as smoking_source_column as data.
  idx <- idx[!grepl("source_column|source_file|_source$", cn[idx], ignore.case = TRUE)]
  if (!length(idx)) return(NULL)

  scores <- vapply(idx, function(i) {
    vals <- clean_missing_strings(df[[i]])
    sum(!is.na(vals))
  }, integer(1))

  cn[idx[which.max(scores)]]
}

standardize_taxon_label <- function(x) {
  x <- as.character(x)
  for (old in names(TAXON_SYNONYMS)) {
    x <- stringr::str_replace_all(x, fixed(old), TAXON_SYNONYMS[[old]])
  }
  x
}

save_table <- function(df, filename) {
  path <- file.path(TABLES_DIR, filename)
  readr::write_csv(df, path, na = "")
  message("Wrote table: ", path)
  invisible(path)
}

safe_figure_dimension <- function(x, fallback, lower = 2, upper = 20) {
  x <- suppressWarnings(as.numeric(x)[1])
  if (!is.finite(x) || x <= 0) x <- fallback
  min(max(x, lower), upper)
}

save_figure <- function(p, filename, width = 7, height = 5, out_dir = FIGURES_DIR) {
  if (is.null(p)) return(invisible(NULL))
  width <- safe_figure_dimension(width, fallback = 7)
  height <- safe_figure_dimension(height, fallback = 5)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  pdf <- file.path(out_dir, paste0(filename, ".pdf"))
  svg <- file.path(out_dir, paste0(filename, ".svg"))

  tryCatch(
    ggplot2::ggsave(pdf, p, width = width, height = height, device = PDF_DEVICE, limitsize = FALSE),
    error = function(e) {
      diagnostic_rds <- file.path(LOG_DIR, paste0(filename, "_failed_plot.rds"))
      diagnostic_txt <- file.path(LOG_DIR, paste0(filename, "_figure_error.txt"))
      saveRDS(p, diagnostic_rds)
      writeLines(
        c(
          paste0("Figure: ", filename),
          paste0("Width: ", width),
          paste0("Height: ", height),
          paste0("Error: ", conditionMessage(e))
        ),
        diagnostic_txt
      )
      stop("Failed to write PDF figure '", filename, "': ", conditionMessage(e), call. = FALSE)
    }
  )

  # SVG is useful for editing but should not invalidate a successful PDF run.
  tryCatch(
    ggplot2::ggsave(svg, p, width = width, height = height, limitsize = FALSE),
    error = function(e) warning("PDF written, but SVG export failed for ", filename, ": ", conditionMessage(e), call. = FALSE)
  )

  message("Wrote figure: ", pdf)
  invisible(list(pdf = pdf, svg = if (file.exists(svg)) svg else NA_character_))
}

fig_theme <- function(base_size = 9, legend_position = "bottom") {
  theme_classic(base_size = base_size) +
    theme(
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      legend.position = legend_position,
      legend.title = element_text(face = "bold"),
      strip.background = element_rect(fill = "grey95", colour = "grey80"),
      strip.text = element_text(face = "bold"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = base_size),
      axis.text = element_text(size = base_size - 1),
      plot.margin = margin(5.5, 14, 5.5, 5.5)
    )
}
```

# 2. Data inventory

This section checks that all expected input files are present and records their basic structure. It is the first chunk to run when debugging paths, file formats, or sample-ID mismatches.

```{r expected-files}
input_files <- unique(c(
  cohort_config$community$metadata_file,
  cohort_config$community$preference_file,
  cohort_config$community$feature_tables$family,
  cohort_config$community$feature_tables$genus,
  cohort_config$student$metadata_file,
  cohort_config$student$feature_tables$family,
  cohort_config$student$feature_tables$genus,
  cohort_config$community$alpha_table,
  cohort_config$student$alpha_table,
  if (RUN_SPECIES_MAASLIN) cohort_config$community$feature_tables$species else NULL,
  if (RUN_FB_RATIO) c(
    cohort_config$community$rarefied_tables$phylum,
    cohort_config$student$rarefied_tables$phylum
  ) else NULL,
  if (RUN_METABOLOMICS) cohort_config$community$metabolome_file else NULL
))
```

```{r inspect-input-functions}
inspect_file <- function(file) {
  path <- file.path(DATA_DIR, file)
  exists <- file.exists(path)
  size_mb <- if (exists) unname(file.info(path)$size / 1024^2) else NA_real_

  make_row <- function(object = NA_character_, class = NA_character_,
                       n_rows = NA_integer_, n_cols = NA_integer_,
                       first_columns = NA_character_, first_rownames = NA_character_,
                       first_colnames = NA_character_, note = NA_character_) {
    tibble(
      file = file,
      exists = exists,
      size_mb = size_mb,
      object = object,
      class = class,
      n_rows = n_rows,
      n_cols = n_cols,
      first_columns = first_columns,
      first_rownames = first_rownames,
      first_colnames = first_colnames,
      note = note
    )
  }

  if (!exists) return(make_row(note = "File not found"))

  rdata_inputs <- unique(stats::na.omit(c(
    cohort_config$community$metadata_file,
    cohort_config$community$metabolome_file
  )))
  if (file %in% rdata_inputs) {
    e <- new.env(parent = emptyenv())
    load(path, envir = e)
    return(purrr::map_dfr(ls(e), function(obj) {
      x <- get(obj, envir = e)
      d <- dim(x)
      make_row(
        object = obj,
        class = paste(class(x), collapse = "; "),
        n_rows = if (!is.null(d)) as.integer(d[1]) else NA_integer_,
        n_cols = if (!is.null(d)) as.integer(d[2]) else NA_integer_,
        first_columns = if (!is.null(colnames(x))) paste(head(colnames(x), 8), collapse = " | ") else NA_character_,
        first_rownames = if (!is.null(rownames(x))) paste(head(rownames(x), 8), collapse = " | ") else NA_character_,
        first_colnames = if (!is.null(colnames(x))) paste(head(colnames(x), 8), collapse = " | ") else NA_character_
      )
    }))
  }

  if (grepl("\\.xlsx$|\\.xls$", file, ignore.case = TRUE)) {
    return(purrr::map_dfr(readxl::excel_sheets(path), function(sh) {
      x <- readxl::read_excel(path, sheet = sh, .name_repair = "minimal")
      make_row(
        object = paste0("sheet:", sh),
        class = paste(class(x), collapse = "; "),
        n_rows = nrow(x),
        n_cols = ncol(x),
        first_columns = paste(head(names(x), 8), collapse = " | "),
        first_rownames = NA_character_,
        first_colnames = paste(head(names(x), 8), collapse = " | ")
      )
    }))
  }

  # The functional annotation table is several gigabytes. Reading it in full
  # solely for the inventory wastes memory; analysis chunks can still load it
  # explicitly if it is ever used.
  if (is.finite(size_mb) && size_mb > 500) {
    x <- tryCatch(
      data.table::fread(path, nrows = 2, data.table = FALSE, check.names = FALSE, showProgress = FALSE),
      error = function(e) NULL
    )
    if (is.null(x)) return(make_row(object = "unreadable", note = "Large-file header read failed"))
    return(make_row(
      object = "large table (header only)",
      class = paste(class(x), collapse = "; "),
      n_rows = NA_integer_,
      n_cols = ncol(x),
      first_columns = paste(head(names(x), 8), collapse = " | "),
      first_rownames = NA_character_,
      first_colnames = paste(head(names(x), 8), collapse = " | "),
      note = "Row count intentionally not evaluated in inventory"
    ))
  }

  x <- tryCatch(
    data.table::fread(path, data.table = FALSE, check.names = FALSE, showProgress = FALSE),
    error = function(e) NULL
  )
  if (is.null(x)) return(make_row(object = "unreadable", note = "fread failed"))

  make_row(
    object = "table",
    class = paste(class(x), collapse = "; "),
    n_rows = nrow(x),
    n_cols = ncol(x),
    first_columns = paste(head(names(x), 8), collapse = " | "),
    first_rownames = NA_character_,
    first_colnames = paste(head(names(x), 8), collapse = " | ")
  )
}
```

```{r input-inventory}
input_inventory <- purrr::map_dfr(input_files, inspect_file)
knitr::kable(input_inventory)
save_table(input_inventory, "input_inventory.csv")
```

# 3. Metadata loading and harmonization

```{r metadata-functions}
load_community_meta <- function() {
  meta_file <- file.path(DATA_DIR, cohort_config$community$metadata_file)
  pref_file <- file.path(DATA_DIR, cohort_config$community$preference_file)

  e <- new.env(parent = emptyenv())
  load(meta_file, envir = e)
  if (!exists("meta_dummy", envir = e, inherits = FALSE)) {
    stop("Expected object 'meta_dummy' in ", meta_file)
  }

  meta <- as.data.frame(get("meta_dummy", envir = e), check.names = FALSE)
  names(meta) <- canon(names(meta))
  meta$sample_id <- coerce_id(rownames(meta))

  prefs <- read_any(pref_file)
  names(prefs) <- canon(names(prefs))
  id_col <- find_col(prefs, c("novoge_sample_id", "sample_id", "sampleid"))
  if (is.null(id_col)) stop("Could not find sample ID column in Hamburg preference file.")
  prefs$sample_id <- coerce_id(prefs[[id_col]])

  meta <- dplyr::left_join(meta, prefs, by = "sample_id", suffix = c("", "_pref"))
  meta <- meta %>%
    dplyr::filter(!is.na(sample_id), sample_id != "") %>%
    dplyr::distinct(sample_id, .keep_all = TRUE)
  rownames(meta) <- meta$sample_id
  meta
}

load_student_meta <- function() {
  meta_file <- file.path(DATA_DIR, cohort_config$student$metadata_file)
  meta <- tryCatch(read_any(meta_file, sheet = "UHHdata"), error = function(e) read_any(meta_file, sheet = 1))
  names(meta) <- canon(names(meta))
  id_col <- find_col(meta, c("kitid", "sample_id", "sampleid"))
  if (is.null(id_col)) stop("Could not find kitid/sample_id in student metadata.")
  meta$sample_id <- coerce_id(meta[[id_col]])
  meta <- meta %>%
    dplyr::filter(!is.na(sample_id), sample_id != "") %>%
    dplyr::distinct(sample_id, .keep_all = TRUE)
  rownames(meta) <- meta$sample_id
  meta
}

harmonize_bmi <- function(meta) {
  bmi_col <- find_col(meta, c("bmi", "body_mass_index", "bodymassindex"))
  weight_col <- find_col(meta, c("weight_kg", "weight", "body_weight", "bodyweight"))
  height_col <- find_col(meta, c("height_cm", "height", "body_height", "bodyheight"))

  raw_bmi <- if (!is.null(bmi_col)) numify(meta[[bmi_col]]) else rep(NA_real_, nrow(meta))
  calc_bmi <- rep(NA_real_, nrow(meta))

  if (!is.null(weight_col) && !is.null(height_col)) {
    w <- numify(meta[[weight_col]])
    h <- numify(meta[[height_col]])
    calc_bmi <- w / (h / 100)^2
    calc_bmi[!is.finite(calc_bmi)] <- NA_real_
  }

  bmi <- raw_bmi
  bad <- is.na(bmi) | bmi < 10 | bmi > 60
  bmi[bad & !is.na(calc_bmi)] <- calc_bmi[bad & !is.na(calc_bmi)]

  meta$bmi_raw <- raw_bmi
  meta$bmi <- bmi
  meta
}

clean_missing_strings <- function(x) {
  x <- as.character(x)
  x <- stringr::str_trim(x)
  x[tolower(x) %in% c("", "na", "n/a", "nan", "null", "missing", "not applicable", "not_applicable", "not sure", "unknown")] <- NA_character_
  x
}

make_smoking_factor <- function(x) {
  smoke_chr <- clean_missing_strings(x)
  observed <- unique(stats::na.omit(smoke_chr))
  preferred_order <- c(
    "Never", "Rarely (less than once/week)", "Rarely (a few times/month)",
    "Occasionally (1-2 times/week)", "Regularly (3-5 times/week)", "Daily"
  )
  if ("Never" %in% observed) {
    lev <- c(preferred_order[preferred_order %in% observed], sort(setdiff(observed, preferred_order)))
    factor(smoke_chr, levels = lev)
  } else {
    factor(smoke_chr)
  }
}

make_alcohol_frequency_factor <- function(x) {
  alcohol_chr <- clean_missing_strings(x)
  observed <- unique(stats::na.omit(alcohol_chr))
  preferred_order <- c(
    "Never", "Rarely (a few times/month)", "Occasionally (1-2 times/week)",
    "Regularly (3-5 times/week)", "Daily"
  )
  if ("Never" %in% observed) {
    lev <- c(preferred_order[preferred_order %in% observed], sort(setdiff(observed, preferred_order)))
    factor(alcohol_chr, levels = lev)
  } else {
    factor(alcohol_chr)
  }
}

make_clubmember_factor <- function(x) {
  factor(clean_missing_strings(x))
}

add_preference_zscores <- function(meta, preference_candidates = PREFERENCE_CANDIDATES) {
  present <- preference_candidates[preference_candidates %in% canon(names(meta))]
  for (p in present) {
    src <- names(meta)[match(p, canon(names(meta)))]
    meta[[paste0("z_", p)]] <- zscore(meta[[src]])
  }
  meta
}

harmonize_model_covariates <- function(meta) {
  # Keep covariate names and classes explicit and reproducible.
  # Especially important: the Student cohort workbook may contain smoking_frequency;
  # ALIASES maps this to smoking, and here smoking is forced to be a factor.

  if ("age" %in% names(meta)) meta$age <- numify(meta$age)
  if ("bmi" %in% names(meta)) meta$bmi <- numify(meta$bmi)

  if ("gender" %in% names(meta)) {
    gender_chr <- clean_missing_strings(meta$gender)
    meta$gender <- factor(gender_chr)
  }

  smoke_col <- find_best_col(
    meta,
    c("smoking", "smoking_frequency", "smoke", "smoking_status", "tobacco", "cigarette"),
    pattern = "smok|tobacco|cigarette"
  )
  if (!is.null(smoke_col)) {
    meta$smoking_source_column <- smoke_col
    meta$smoking <- make_smoking_factor(meta[[smoke_col]])
  } else {
    meta$smoking_source_column <- NA_character_
    meta$smoking <- factor(rep(NA_character_, nrow(meta)))
  }

  if ("clubmember" %in% names(meta)) {
    meta$clubmember <- make_clubmember_factor(meta$clubmember)
  }

  if ("alcohol_frequency" %in% names(meta)) {
    meta$alcohol_frequency <- make_alcohol_frequency_factor(meta$alcohol_frequency)
  }

  meta
}

add_ecological_quartiles <- function(meta) {
  labels_q <- c("Q1 lowest", "Q2", "Q3", "Q4 highest")

  if (!"z_ecological" %in% names(meta)) {
    meta$ecological_preference_q <- factor("Missing", levels = c(labels_q, "Missing"))
    return(meta)
  }

  z <- meta$z_ecological
  q <- rep(NA_integer_, length(z))
  ok <- is.finite(z)
  if (sum(ok) >= 4) q[ok] <- dplyr::ntile(z[ok], 4)

  lab <- rep("Missing", length(z))
  lab[!is.na(q)] <- labels_q[q[!is.na(q)]]
  meta$ecological_preference_q <- factor(lab, levels = c(labels_q, "Missing"))
  meta
}

prepare_metadata_for_models <- function(meta) {
  meta %>%
    harmonize_bmi() %>%
    harmonize_model_covariates() %>%
    add_preference_zscores() %>%
    add_ecological_quartiles()
}
```

```{r load-metadata}
meta_community <- load_community_meta() %>% prepare_metadata_for_models()
meta_student <- load_student_meta() %>% prepare_metadata_for_models()

metadata_summary <- tibble(
  cohort = c("Community cohort", "Student cohort"),
  n = c(nrow(meta_community), nrow(meta_student)),
  n_preferences = c(
    sum(startsWith(names(meta_community), "z_")),
    sum(startsWith(names(meta_student), "z_"))
  ),
  n_bmi_missing = c(sum(is.na(meta_community$bmi)), sum(is.na(meta_student$bmi))),
  n_smoking_missing = c(sum(is.na(meta_community$smoking)), sum(is.na(meta_student$smoking))),
  smoking_levels = c(
    paste(levels(droplevels(meta_community$smoking)), collapse = " | "),
    paste(levels(droplevels(meta_student$smoking)), collapse = " | ")
  )
)

knitr::kable(metadata_summary)
save_table(metadata_summary, "metadata_summary.csv")

saveRDS(meta_community, file.path(RESULTS_DIR, "metadata_community.rds"))
saveRDS(meta_student, file.path(RESULTS_DIR, "metadata_student.rds"))
readr::write_csv(meta_community, file.path(TABLES_DIR, "metadata_community.csv"))
readr::write_csv(meta_student, file.path(TABLES_DIR, "metadata_student.csv"))
```

## 3.0a Community preference-linkage audit

This audit documents the linkage between the 138 Community microbiome records and
`Hamburg_AltruismVariables.csv`. It is intentionally run from the two source files
rather than from the already joined object so that ID-normalization problems are
visible. With the corrected one-digit ID handling, all 138 IDs should overlap.

```{r community-preference-linkage-audit}
community_meta_source_env <- new.env(parent = emptyenv())
load(
  file.path(DATA_DIR, cohort_config$community$metadata_file),
  envir = community_meta_source_env
)

community_meta_source <- as.data.frame(
  get("meta_dummy", envir = community_meta_source_env),
  check.names = FALSE
)
names(community_meta_source) <- canon(names(community_meta_source))
community_meta_source$sample_id <- coerce_id(rownames(community_meta_source))

community_pref_source <- read_any(
  file.path(DATA_DIR, cohort_config$community$preference_file)
)
names(community_pref_source) <- canon(names(community_pref_source))
community_pref_id_col <- find_col(
  community_pref_source,
  c("novoge_sample_id", "sample_id", "sampleid")
)
if (is.null(community_pref_id_col)) {
  stop("Could not identify the Community preference sample-ID column.")
}
community_pref_source$sample_id <- coerce_id(
  community_pref_source[[community_pref_id_col]]
)

community_preference_linkage_audit <- tibble(
  dataset = c(
    "meta_dummy",
    "Hamburg_AltruismVariables.csv",
    "harmonized meta_community",
    "IDs shared by meta_dummy and preference file"
  ),
  n = c(
    dplyr::n_distinct(community_meta_source$sample_id),
    dplyr::n_distinct(community_pref_source$sample_id),
    dplyr::n_distinct(rownames(meta_community)),
    length(intersect(
      community_meta_source$sample_id,
      community_pref_source$sample_id
    ))
  )
)

community_preference_unmatched_ids <- dplyr::bind_rows(
  tibble(
    issue = "preference_id_not_in_meta_dummy",
    sample_id = setdiff(
      community_pref_source$sample_id,
      community_meta_source$sample_id
    )
  ),
  tibble(
    issue = "meta_dummy_id_without_preference_record",
    sample_id = setdiff(
      community_meta_source$sample_id,
      community_pref_source$sample_id
    )
  )
)

knitr::kable(community_preference_linkage_audit)
knitr::kable(community_preference_unmatched_ids)
save_table(
  community_preference_linkage_audit,
  "community_preference_linkage_audit.csv"
)
save_table(
  community_preference_unmatched_ids,
  "community_preference_unmatched_ids.csv"
)
```

# 3.1 Cohort-description table

This section generates the descriptive cohort characteristics used for manuscript
Table 1 directly from the harmonized analysis metadata. Preference medians are
reported on the original questionnaire scale (0--10). The source variables are
stored as 1--11, so an offset of 1 is removed for descriptive reporting only.
Donation remains on its original 0--50 EUR scale. The inferential models are not
changed by this reporting transformation because preference predictors are
z-standardized within cohort.

```{r cohort-description}
female_indicator <- function(x, female_code = "1") {
  x_chr <- clean_missing_strings(x)
  out <- ifelse(is.na(x_chr), NA_real_, as.numeric(x_chr == female_code))
  as.numeric(out)
}

get_numeric_column <- function(meta, candidates) {
  col <- find_col(meta, candidates)
  if (is.null(col)) return(rep(NA_real_, nrow(meta)))
  numify(meta[[col]])
}

get_preference_raw <- function(meta, var) {
  col <- find_col(meta, c(var))
  if (is.null(col)) return(rep(NA_real_, nrow(meta)))
  numify(meta[[col]])
}

preference_for_description <- function(meta, var, cohort_key) {
  x <- get_preference_raw(meta, var)
  offset <- unname(PREFERENCE_DESCRIPTIVE_OFFSET[[cohort_key]])
  if (!is.finite(offset)) offset <- 0
  x - offset
}

make_cohort_description_long <- function(meta, cohort_key, cohort_label, female_code) {
  age <- get_numeric_column(meta, c("age", "age_years"))
  bmi <- get_numeric_column(meta, c("bmi", "body_mass_index", "bodymassindex"))
  weight <- get_numeric_column(meta, c("weight_kg", "weight", "body_weight", "bodyweight"))
  female <- if ("gender" %in% names(meta)) {
    female_indicator(meta$gender, female_code)
  } else {
    rep(NA_real_, nrow(meta))
  }

  donation <- get_preference_raw(meta, "donation")
  risk <- preference_for_description(meta, "risktaking", cohort_key)
  patience <- preference_for_description(meta, "timediscounting", cohort_key)
  trust <- preference_for_description(meta, "trust", cohort_key)
  selfcontrol <- preference_for_description(meta, "selfcontrol", cohort_key)
  ecological <- preference_for_description(meta, "ecological", cohort_key)

  tibble(
    cohort = cohort_label,
    sample_id = rownames(meta),
    age = age,
    female = female,
    bmi = bmi,
    weight = weight,
    donation = donation,
    risk_taking = risk,
    patience = patience,
    trust = trust,
    self_control = selfcontrol,
    ecological_orientation = ecological
  )
}

cohort_description_data <- dplyr::bind_rows(
  make_cohort_description_long(
    meta_community,
    "community",
    "Community cohort",
    FEMALE_CODE[["community"]]
  ),
  make_cohort_description_long(
    meta_student,
    "student",
    "Student cohort",
    FEMALE_CODE[["student"]]
  )
)

cohort_description <- cohort_description_data %>%
  dplyr::group_by(cohort) %>%
  dplyr::summarise(
    n = dplyr::n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = stats::sd(age, na.rm = TRUE),
    age_n = sum(is.finite(age)),
    female_percent = 100 * mean(female, na.rm = TRUE),
    gender_n = sum(!is.na(female)),
    bmi_mean = mean(bmi, na.rm = TRUE),
    bmi_sd = stats::sd(bmi, na.rm = TRUE),
    bmi_n = sum(is.finite(bmi)),
    weight_mean = mean(weight, na.rm = TRUE),
    weight_sd = stats::sd(weight, na.rm = TRUE),
    weight_n = sum(is.finite(weight)),
    donation_mean = mean(donation, na.rm = TRUE),
    donation_sd = stats::sd(donation, na.rm = TRUE),
    donation_n = sum(is.finite(donation)),
    donors_percent = 100 * mean(donation > 0, na.rm = TRUE),
    risk_taking_median = median(risk_taking, na.rm = TRUE),
    patience_median = median(patience, na.rm = TRUE),
    trust_median = median(trust, na.rm = TRUE),
    self_control_median = median(self_control, na.rm = TRUE),
    ecological_orientation_median = median(ecological_orientation, na.rm = TRUE),
    .groups = "drop"
  )

# Auditable long-form summaries with n/min/median/max for each preference.
cohort_preference_descriptive_audit <- cohort_description_data %>%
  dplyr::select(
    cohort, sample_id, donation, risk_taking, patience, trust,
    self_control, ecological_orientation
  ) %>%
  tidyr::pivot_longer(
    -c(cohort, sample_id),
    names_to = "variable",
    values_to = "value"
  ) %>%
  dplyr::group_by(cohort, variable) %>%
  dplyr::summarise(
    n = sum(is.finite(value)),
    min = if (any(is.finite(value))) min(value, na.rm = TRUE) else NA_real_,
    mean = if (any(is.finite(value))) mean(value, na.rm = TRUE) else NA_real_,
    sd = if (sum(is.finite(value)) > 1) stats::sd(value, na.rm = TRUE) else NA_real_,
    median = if (any(is.finite(value))) median(value, na.rm = TRUE) else NA_real_,
    max = if (any(is.finite(value))) max(value, na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )

# Manuscript-ready table values in the same order as Table 1.
cohort_description_manuscript <- cohort_description %>%
  dplyr::transmute(
    cohort,
    `Sample size` = n,
    `Age, mean (SD), years` = sprintf("%.1f (%.1f)", age_mean, age_sd),
    `Female participants, %` = sprintf("%.1f", female_percent),
    `BMI, mean (SD), kg/m2` = sprintf("%.1f (%.1f)", bmi_mean, bmi_sd),
    `Weight, mean (SD), kg` = ifelse(
      is.finite(weight_mean),
      sprintf("%.1f (%.1f)", weight_mean, weight_sd),
      NA_character_
    ),
    `Donation (in Euro), mean (SD)` = sprintf(
      "%.1f (%.1f)", donation_mean, donation_sd
    ),
    `Donors, %` = sprintf("%.1f", donors_percent),
    `Risk-taking, median` = sprintf("%g", risk_taking_median),
    `Patience/time discounting, median` = sprintf("%g", patience_median),
    `Trust, median` = sprintf("%g", trust_median),
    `Self-control, median` = sprintf("%g", self_control_median),
    `Ecological orientation, median` = sprintf("%g", ecological_orientation_median)
  )

knitr::kable(cohort_description, digits = 2)
knitr::kable(cohort_description_manuscript)
knitr::kable(cohort_preference_descriptive_audit, digits = 2)

save_table(cohort_description, "cohort_description.csv")
save_table(
  cohort_description_manuscript,
  "cohort_description_manuscript_table1.csv"
)
save_table(
  cohort_preference_descriptive_audit,
  "cohort_preference_descriptive_audit.csv"
)
```

# 4. Feature-table loading and sample overlap

```{r feature-table-functions}
read_feature_table <- function(path) {
  df <- read_any(path)
  if (is.null(df) || ncol(df) < 2) return(NULL)
  names(df) <- make.unique(names(df))
  tax <- standardize_taxon_label(df[[1]])
  mat <- as.matrix(df[, -1, drop = FALSE])
  suppressWarnings(storage.mode(mat) <- "numeric")
  mat[is.na(mat)] <- 0
  rownames(mat) <- make.unique(as.character(tax))
  colnames(mat) <- coerce_id(names(df)[-1])
  mat
}

align_feature_to_meta <- function(feature_mat, meta) {
  feature_ids <- coerce_id(colnames(feature_mat))
  meta_ids <- rownames(meta)
  common <- intersect(feature_ids, meta_ids)
  if (!length(common)) {
    stop("No sample overlap between feature table and metadata.")
  }
  list(
    feature = feature_mat[, match(common, feature_ids), drop = FALSE],
    meta = meta[match(common, meta_ids), , drop = FALSE],
    common = common
  )
}

# Convert a metadata/data-frame object with sample IDs stored in row names
# into an explicit sample_id column, without duplicating an existing sample_id column.
# This avoids: Error in tibble::rownames_to_column(): Column name `sample_id` must not be duplicated.
meta_with_sample_id <- function(meta, id_col = "sample_id") {
  out <- as.data.frame(meta, check.names = FALSE)
  out[[id_col]] <- NULL
  tibble::rownames_to_column(out, var = id_col)
}
```

```{r load-feature-tables}
feature_paths <- list(
  community_family = file.path(DATA_DIR, cohort_config$community$feature_tables$family),
  community_genus = file.path(DATA_DIR, cohort_config$community$feature_tables$genus),
  community_species = file.path(DATA_DIR, cohort_config$community$feature_tables$species),
  community_phylum = file.path(DATA_DIR, cohort_config$community$rarefied_tables$phylum),
  student_family = file.path(DATA_DIR, cohort_config$student$feature_tables$family),
  student_genus = file.path(DATA_DIR, cohort_config$student$feature_tables$genus),
  student_phylum = file.path(DATA_DIR, cohort_config$student$rarefied_tables$phylum)
)

features <- purrr::map(feature_paths, read_feature_table)
required_feature_names <- c(
  "community_family", "community_genus",
  "student_family", "student_genus"
)
if (RUN_FB_RATIO) {
  required_feature_names <- c(required_feature_names, "community_phylum", "student_phylum")
}
missing_required_features <- required_feature_names[
  purrr::map_lgl(features[required_feature_names], is.null)
]
if (length(missing_required_features)) {
  stop(
    "Required feature table(s) could not be loaded: ",
    paste(missing_required_features, collapse = ", "),
    call. = FALSE
  )
}

sample_overlap <- tibble(
  table = names(features),

  n_taxa = purrr::map_int(features, function(ft) {
    if (is.null(ft)) return(NA_integer_)
    nrow(ft)
  }),

  n_samples_raw = purrr::map_int(features, function(ft) {
    if (is.null(ft)) return(NA_integer_)
    ncol(ft)
  }),

  n_overlap = purrr::imap_int(features, function(ft, nm) {

    if (is.null(ft))
      return(NA_integer_)

    meta <- if (startsWith(nm, "community"))
      meta_community
    else
      meta_student

    length(intersect(
      coerce_id(colnames(ft)),
      rownames(meta)
    ))
  })
)

knitr::kable(sample_overlap)
save_table(sample_overlap, "sample_overlap_feature_tables.csv")

```

# 5. Alpha diversity: one table per cohort

For the manuscript and QC report, alpha diversity is read from exactly **one RTK median alpha-diversity table per cohort**:

- Community cohort: `r cohort_config$community$alpha_table`
- Student cohort: `r cohort_config$student$alpha_table`

This avoids accidentally mixing alpha-diversity estimates derived from different taxonomic resolutions.

```{r alpha-selected-files}
alpha_files_selected <- tibble(
  cohort = c("Community cohort", "Student cohort"),
  file = c(cohort_config$community$alpha_table, cohort_config$student$alpha_table),
  path = file.path(DATA_DIR, file),
  exists = file.exists(path)
)

knitr::kable(alpha_files_selected)
save_table(alpha_files_selected, "alpha_selected_files.csv")

if (any(!alpha_files_selected$exists)) {
  stop(
    "Missing selected alpha-diversity file(s): ",
    paste(alpha_files_selected$file[!alpha_files_selected$exists], collapse = ", ")
  )
}
```

```{r alpha-functions-fixed}
read_rtk_median_alpha <- function(path, cohort_label, meta = NULL, level_label = "RTK median alpha") {
  if (is.null(path) || !file.exists(path)) {
    message("Alpha file not found: ", path)
    return(tibble())
  }

  # RTK median-alpha files can appear in two slightly different forms:
  # 1) depth row + metric-header row + samples
  #      depth  7423 7423 ...
  #      Smpl   Richness Shannon ...
  #      X103... 233 3.20 ...
  # 2) metric-header row + samples
  #      Smpl   Richness Shannon ...
  #      D105   359 4.34 ...
  # Read without column names and then reconstruct the correct header.
  raw <- readr::read_tsv(
    path,
    col_names = FALSE,
    show_col_types = FALSE,
    progress = FALSE,
    name_repair = "minimal"
  )

  raw <- as.data.frame(raw, check.names = FALSE)
  raw <- raw[rowSums(!is.na(raw)) > 0, , drop = FALSE]

  if (nrow(raw) < 2 || ncol(raw) < 3) {
    message("Alpha file too small or malformed: ", path)
    return(tibble())
  }

  first_col <- canon(raw[[1]])

  if (first_col[1] == "depth" && nrow(raw) >= 2 && first_col[2] %in% c("smpl", "sample", "sample_id", "sampleid")) {
    header <- as.character(unlist(raw[2, ], use.names = FALSE))
    dat <- raw[-c(1, 2), , drop = FALSE]
  } else if (first_col[1] %in% c("smpl", "sample", "sample_id", "sampleid")) {
    header <- as.character(unlist(raw[1, ], use.names = FALSE))
    dat <- raw[-1, , drop = FALSE]
  } else {
    # fallback: try normal read with header if the file already had a conventional header
    normal <- read_any(path)
    if (is.null(normal)) return(tibble())
    dat <- normal
    header <- names(dat)
  }

  header <- janitor::make_clean_names(header)
  names(dat) <- make.unique(header)
  names(dat) <- canon(names(dat))

  id_col <- intersect(c("smpl", "sample", "sample_id", "sampleid", "id"), names(dat))[1]
  richness_col <- intersect(c("richness", "observed", "observed_richness", "smpl_richness"), names(dat))[1]
  shannon_col <- intersect(c("shannon", "shannon_diversity"), names(dat))[1]
  simpson_col <- intersect(c("simpson"), names(dat))[1]
  invsimpson_col <- intersect(c("inv_simpson", "inverse_simpson", "inv_simpson_1"), names(dat))[1]
  chao1_col <- intersect(c("chao1", "chao"), names(dat))[1]
  evenness_col <- intersect(c("eveness", "evenness", "pielou", "pielou_evenness"), names(dat))[1]

  if (is.na(id_col) || is.na(richness_col) || is.na(shannon_col)) {
    message("Could not identify required alpha columns in: ", path)
    message("Columns are: ", paste(names(dat), collapse = ", "))
    return(tibble())
  }

  out <- tibble(
    sample_id_original = as.character(dat[[id_col]]),
    sample_id = coerce_id(dat[[id_col]]),
    cohort = cohort_label,
    level = level_label,
    richness = numify(dat[[richness_col]]),
    shannon = numify(dat[[shannon_col]]),
    simpson = if (!is.na(simpson_col)) numify(dat[[simpson_col]]) else NA_real_,
    inv_simpson = if (!is.na(invsimpson_col)) numify(dat[[invsimpson_col]]) else NA_real_,
    chao1 = if (!is.na(chao1_col)) numify(dat[[chao1_col]]) else NA_real_,
    evenness = if (!is.na(evenness_col)) numify(dat[[evenness_col]]) else NA_real_,
    source_file = basename(path)
  ) %>%
    dplyr::filter(is.finite(richness), is.finite(shannon))

  if (!is.null(meta)) {
    before <- nrow(out)
    out <- out %>% dplyr::filter(sample_id %in% rownames(meta))
    message(
      cohort_label, " alpha: ", nrow(out), "/", before,
      " rows overlapped with metadata from ", basename(path)
    )
  }

  out
}

plot_alpha_cohort <- function(alpha_df, filename, width = 6.4, height = 3.8) {
  if (!nrow(alpha_df)) stop("No alpha-diversity rows available for plotting.")

  alpha_long <- alpha_df %>%
    tidyr::pivot_longer(c(shannon, richness), names_to = "metric", values_to = "value") %>%
    dplyr::mutate(
      cohort = factor(cohort, levels = c("Community cohort", "Student cohort")),
      metric = factor(metric,
                      levels = c("shannon", "richness"),
                      labels = c("Shannon diversity", "Observed richness"))
    ) %>%
    dplyr::filter(is.finite(value))

  stats <- alpha_long %>%
    dplyr::group_by(metric) %>%
    dplyr::summarise(
      n_community = sum(cohort == "Community cohort"),
      n_student = sum(cohort == "Student cohort"),
      p_value = if (dplyr::n_distinct(cohort) == 2 && min(table(cohort)) > 0) {
        tryCatch(stats::wilcox.test(value ~ cohort)$p.value, error = function(e) NA_real_)
      } else NA_real_,
      p_label = dplyr::case_when(
        is.na(p_value) ~ "",
        p_value < 0.001 ~ "P < 0.001",
        TRUE ~ paste0("P = ", signif(p_value, 2))
      ),
      y = max(value, na.rm = TRUE) * 1.08,
      .groups = "drop"
    )

  p <- ggplot(alpha_long, aes(x = cohort, y = value, fill = cohort)) +
    geom_violin(trim = FALSE, alpha = 0.45, colour = NA, width = 0.85) +
    geom_boxplot(width = 0.14, outlier.shape = NA, alpha = 0.85, linewidth = 0.3) +
    geom_point(position = position_jitter(width = 0.08, height = 0), alpha = 0.35, size = 0.65, show.legend = FALSE) +
    geom_text(data = stats, aes(x = 1.5, y = y, label = p_label), inherit.aes = FALSE, size = 2.6) +
    facet_wrap(~ metric, scales = "free_y", nrow = 1) +
    scale_fill_manual(values = COHORT_COLORS, drop = FALSE) +
    labs(x = NULL, y = NULL) +
    fig_theme(base_size = 9, legend_position = "none")

  save_figure(p, filename, width = width, height = height)
  save_table(alpha_long, paste0(filename, "_data.csv"))
  save_table(stats, paste0(filename, "_stats.csv"))
  p
}
```

```{r alpha-main-one-file-per-cohort}
alpha_community <- read_rtk_median_alpha(
  path = file.path(DATA_DIR, cohort_config$community$alpha_table),
  cohort_label = "Community cohort",
  meta = meta_community,
  level_label = "Selected RTK alpha table"
)

alpha_student <- read_rtk_median_alpha(
  path = file.path(DATA_DIR, cohort_config$student$alpha_table),
  cohort_label = "Student cohort",
  meta = meta_student,
  level_label = "Selected RTK alpha table"
)

alpha_main <- bind_rows(alpha_community, alpha_student)

alpha_source_summary <- alpha_main %>%
  dplyr::count(cohort, source_file, name = "n_samples")

knitr::kable(alpha_source_summary)
save_table(alpha_source_summary, "alpha_source_summary.csv")

alpha_ids_community <- unique(alpha_main$sample_id[alpha_main$cohort == "Community cohort"])
alpha_ids_student <- unique(alpha_main$sample_id[alpha_main$cohort == "Student cohort"])

alpha_sample_diagnostics <- tibble(
  cohort = c("Community cohort", "Student cohort"),
  metadata_n = c(nrow(meta_community), nrow(meta_student)),
  alpha_n = c(length(alpha_ids_community), length(alpha_ids_student)),
  overlap_n = c(
    length(intersect(rownames(meta_community), alpha_ids_community)),
    length(intersect(rownames(meta_student), alpha_ids_student))
  ),
  metadata_not_in_alpha_n = c(
    length(setdiff(rownames(meta_community), alpha_ids_community)),
    length(setdiff(rownames(meta_student), alpha_ids_student))
  ),
  alpha_not_in_metadata_n = c(
    length(setdiff(alpha_ids_community, rownames(meta_community))),
    length(setdiff(alpha_ids_student, rownames(meta_student)))
  )
)

knitr::kable(alpha_sample_diagnostics)
save_table(alpha_sample_diagnostics, "alpha_sample_diagnostics.csv")

alpha_missing_samples <- bind_rows(
  tibble(cohort = "Community cohort", issue = "metadata_not_in_alpha",
         sample_id = setdiff(rownames(meta_community), alpha_ids_community)),
  tibble(cohort = "Community cohort", issue = "alpha_not_in_metadata",
         sample_id = setdiff(alpha_ids_community, rownames(meta_community))),
  tibble(cohort = "Student cohort", issue = "metadata_not_in_alpha",
         sample_id = setdiff(rownames(meta_student), alpha_ids_student)),
  tibble(cohort = "Student cohort", issue = "alpha_not_in_metadata",
         sample_id = setdiff(alpha_ids_student, rownames(meta_student)))
)
save_table(alpha_missing_samples, "alpha_missing_samples.csv")

alpha_summary <- alpha_main %>%
  tidyr::pivot_longer(c(shannon, richness, simpson, inv_simpson, chao1, evenness),
                      names_to = "metric", values_to = "value") %>%
  dplyr::filter(is.finite(value)) %>%
  dplyr::group_by(cohort, metric) %>%
  dplyr::summarise(
    n = dplyr::n(),
    median = median(value, na.rm = TRUE),
    iqr_low = quantile(value, 0.25, na.rm = TRUE),
    iqr_high = quantile(value, 0.75, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(alpha_summary)
save_table(alpha_summary, "alpha_main_summary.csv")

p_alpha_main <- plot_alpha_cohort(alpha_main, "alpha_diversity_selected_tables_publication")
p_alpha_main
```

**Alpha-diversity interpretation draft.**

```{r alpha-interpretation, results='asis'}
cat(
  "Alpha diversity was read from one selected RTK median alpha-diversity table per cohort: ",
  "`", cohort_config$community$alpha_table, "` for the community cohort and `",
  cohort_config$student$alpha_table, "` for the student cohort. ",
  "These cohort-specific inputs avoid mixing RTK outputs from different taxonomic resolutions.\n",
  sep = ""
)
```

# 6. Beta diversity, PCoA, and multivariable taxon models

Stage 1 models include all standardized preference terms plus age, gender, and BMI. Stage 2 adds cohort-specific lifestyle covariates. In the Student cohort, smoking, club membership, and alcohol-consumption frequency are categorical Stage 2 covariates; `not applicable` responses are treated as missing where applicable, and `Never` is the reference level for smoking and alcohol frequency when present. Raw religiousness is not added again as a Community-cohort Stage 2 covariate because standardized religiousness is already a preference exposure in the same model.

```{r model-functions}
model_value <- function(x, varname = NULL) {
  if (is.factor(x)) return(droplevels(x))
  if (is.character(x)) return(factor(clean_missing_strings(x)))
  if (!is.null(varname) && varname %in% c("gender", "smoking", "marital_status", "clubmember", "alcohol_frequency")) {
    return(factor(clean_missing_strings(x)))
  }
  numify(x)
}

drop_unused_factor_levels <- function(df) {
  out <- as.data.frame(df, check.names = FALSE)
  for (nm in names(out)) {
    if (is.factor(out[[nm]])) out[[nm]] <- droplevels(out[[nm]])
  }
  out
}

preference_terms <- function(meta) {
  intersect(paste0("z_", PREFERENCE_CANDIDATES), names(meta))
}

stage_covariates <- function(meta, cohort_key, stage = c("stage1", "stage2")) {
  stage <- match.arg(stage)
  cfg <- cohort_config[[cohort_key]]
  core <- cfg$stage1
  extra <- if (stage == "stage2") cfg$stage2 else character(0)
  vars <- unique(c(core, extra))
  vars[vars %in% names(meta)]
}

build_model_design <- function(meta, cohort_key, stage = c("stage1", "stage2"), include_preferences = TRUE) {
  stage <- match.arg(stage)
  meta <- as.data.frame(meta, check.names = FALSE)

  # Re-harmonize Student Stage 2 categorical covariates defensively at
  # model-building time so all downstream model families use identical coding.
  if (cohort_key == "student" && stage == "stage2") {
    if ("smoking" %in% names(meta) && length(unique(stats::na.omit(meta$smoking))) > 1) {
      meta$smoking <- make_smoking_factor(meta$smoking)
    } else {
      smoke_col <- find_best_col(
        meta,
        c("smoking", "smoking_frequency", "smoke", "smoking_status", "tobacco", "cigarette"),
        pattern = "smok|tobacco|cigarette"
      )
      if (!is.null(smoke_col)) meta$smoking <- make_smoking_factor(meta[[smoke_col]])
    }

    if ("clubmember" %in% names(meta)) {
      meta$clubmember <- make_clubmember_factor(meta$clubmember)
    }

    if ("alcohol_frequency" %in% names(meta)) {
      meta$alcohol_frequency <- make_alcohol_frequency_factor(meta$alcohol_frequency)
    }
  }

  required_covars <- stage_covariates(meta, cohort_key, stage)
  optional_preferences <- if (include_preferences) preference_terms(meta) else character(0)
  vars <- unique(c(required_covars, optional_preferences))
  vars <- vars[vars %in% names(meta)]

  design <- meta[, vars, drop = FALSE]
  for (v in names(design)) design[[v]] <- model_value(design[[v]], v)

  required_covars <- intersect(required_covars, names(design))
  informative <- vapply(design, function(x) length(unique(stats::na.omit(x))) > 1, logical(1))

  required_bad <- required_covars[!informative[required_covars]]
  if (length(required_bad)) {
    stop(
      "Required covariate(s) lack variation or are all missing: ",
      paste(required_bad, collapse = ", "),
      call. = FALSE
    )
  }

  optional_terms <- setdiff(names(design), required_covars)
  optional_terms <- optional_terms[informative[optional_terms]]
  design <- design[, c(required_covars, optional_terms), drop = FALSE]

  # Protect required covariates and drop only non-estimable optional terms.
  if (ncol(design)) {
    cc <- stats::complete.cases(design)
    design_cc <- drop_unused_factor_levels(design[cc, , drop = FALSE])

    if (nrow(design_cc) > 2) {
      keep_terms <- character(0)
      for (v in required_covars) {
        trial <- c(keep_terms, v)
        mm <- tryCatch(stats::model.matrix(stats::reformulate(trial), data = design_cc), error = function(e) NULL)
        if (is.null(mm) || qr(mm)$rank < ncol(mm)) {
          stop(
            "Required covariate '", v,
            "' is not estimable after complete-case filtering.",
            call. = FALSE
          )
        }
        keep_terms <- trial
      }
      for (v in setdiff(names(design_cc), required_covars)) {
        trial <- c(keep_terms, v)
        mm <- tryCatch(stats::model.matrix(stats::reformulate(trial), data = design_cc), error = function(e) NULL)
        if (!is.null(mm) && qr(mm)$rank == ncol(mm)) keep_terms <- trial
      }
      design <- design[, keep_terms, drop = FALSE]
    }
  }

  if (cohort_key == "student" && stage == "stage2") {
    required_student_stage2 <- LIFESTYLE_CANDIDATES_STUDENT
    missing_stage2 <- setdiff(required_student_stage2, names(design))
    if (length(missing_stage2)) {
      stop(
        "Student cohort Stage 2 design lost required covariate(s): ",
        paste(missing_stage2, collapse = ", "),
        call. = FALSE
      )
    }

    design$smoking <- droplevels(make_smoking_factor(design$smoking))
    design$clubmember <- droplevels(make_clubmember_factor(design$clubmember))
    design$alcohol_frequency <- droplevels(make_alcohol_frequency_factor(design$alcohol_frequency))
  }

  design
}

write_design_audit <- function(design, cohort_key, level_label, stage_label, phase = "final") {
  design <- as.data.frame(design, check.names = FALSE)
  audit <- tibble(
    cohort = cohort_config[[cohort_key]]$cohort,
    level = level_label,
    stage = stage_label,
    phase = phase,
    n_model_samples = nrow(design),
    variable = names(design),
    class = vapply(design, function(x) paste(class(x), collapse = "/"), character(1)),
    n_non_missing = vapply(design, function(x) sum(!is.na(x)), integer(1)),
    n_levels_or_unique = vapply(design, function(x) length(unique(stats::na.omit(x))), integer(1)),
    levels = vapply(design, function(x) {
      if (is.factor(x) || is.character(x)) {
        paste(head(unique(stats::na.omit(as.character(x))), 10), collapse = " | ")
      } else ""
    }, character(1))
  )
  save_table(
    audit,
    paste0("design_audit_", cohort_key, "_", stage_label, "_", tolower(level_label), "_", phase, ".csv")
  )
  audit
}

run_permanova <- function(feature_mat, meta, cohort_key, level_label, stage_label, permutations = PERMUTATIONS) {
  if (is.null(feature_mat)) return(tibble())
  aligned <- align_feature_to_meta(feature_mat, meta)

  comm <- t(aligned$feature)
  suppressWarnings(storage.mode(comm) <- "numeric")
  comm[!is.finite(comm)] <- 0
  comm <- comm[rowSums(comm, na.rm = TRUE) > 0, , drop = FALSE]
  comm <- comm[, colSums(comm, na.rm = TRUE) > 0, drop = FALSE]

  design <- build_model_design(
    aligned$meta[rownames(comm), , drop = FALSE],
    cohort_key = cohort_key,
    stage = stage_label,
    include_preferences = TRUE
  )
  cc <- stats::complete.cases(design)
  design <- drop_unused_factor_levels(design[cc, , drop = FALSE])
  comm <- comm[rownames(design), , drop = FALSE]

  if (nrow(comm) < 5 || ncol(comm) < 2 || ncol(design) < 1) return(tibble())

  rel <- vegan::decostand(comm, method = "total")
  rel[!is.finite(rel)] <- 0
  bray <- vegan::vegdist(rel, method = "bray")
  set.seed(RANDOM_SEED)
  ad <- tryCatch(
    vegan::adonis2(bray ~ ., data = design, permutations = permutations, by = "margin"),
    error = function(e) {
      warning("PERMANOVA failed for ", cohort_key, "/", level_label, "/", stage_label, ": ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(ad)) return(tibble())

  as.data.frame(ad) %>%
    tibble::rownames_to_column("term") %>%
    dplyr::mutate(
      cohort = cohort_config[[cohort_key]]$cohort,
      cohort_key = cohort_key,
      level = level_label,
      stage = stage_label,
      n_samples = nrow(design),
      permutations = permutations,
      model_terms = paste(names(design), collapse = "; ")
    )
}

plot_pcoa_ecological <- function(feature_mat, meta, cohort_label, level_label, filename, width = 6.5, height = 4.8) {
  if (is.null(feature_mat)) return(NULL)
  aligned <- align_feature_to_meta(feature_mat, meta)

  comm <- t(aligned$feature)
  suppressWarnings(storage.mode(comm) <- "numeric")
  comm[!is.finite(comm)] <- 0
  comm <- comm[rowSums(comm, na.rm = TRUE) > 0, , drop = FALSE]
  comm <- comm[, colSums(comm, na.rm = TRUE) > 0, drop = FALSE]
  if (nrow(comm) < 5 || ncol(comm) < 2) return(NULL)

  rel <- vegan::decostand(comm, method = "total")
  rel[!is.finite(rel)] <- 0
  bray <- vegan::vegdist(rel, method = "bray")
  pcoa <- stats::cmdscale(bray, eig = TRUE, k = 2, add = TRUE)

  positive_eig <- pcoa$eig[pcoa$eig > 0]
  explained <- if (length(positive_eig) >= 2) {
    100 * pmax(pcoa$eig[1:2], 0) / sum(positive_eig)
  } else c(NA_real_, NA_real_)

  plot_df <- tibble(
    sample_id = rownames(rel),
    PCoA1 = pcoa$points[, 1],
    PCoA2 = pcoa$points[, 2]
  ) %>%
    dplyr::left_join(
      meta_with_sample_id(aligned$meta) %>%
        dplyr::select(sample_id, ecological_preference_q),
      by = "sample_id"
    ) %>%
    dplyr::mutate(
      ecological_preference_q = as.character(ecological_preference_q),
      ecological_preference_q = dplyr::if_else(
        is.na(ecological_preference_q) | ecological_preference_q == "",
        "Missing",
        ecological_preference_q
      ),
      ecological_preference_q = factor(
        ecological_preference_q,
        levels = names(ECOLOGICAL_QUARTILE_COLORS)
      )
    ) %>%
    dplyr::filter(is.finite(PCoA1), is.finite(PCoA2))

  ellipse_groups <- plot_df %>%
    dplyr::group_by(ecological_preference_q) %>%
    dplyr::summarise(
      n = dplyr::n(),
      n_unique = dplyr::n_distinct(paste(PCoA1, PCoA2)),
      var_x = stats::var(PCoA1),
      var_y = stats::var(PCoA2),
      .groups = "drop"
    ) %>%
    dplyr::filter(n >= 4, n_unique >= 4, is.finite(var_x), var_x > 0, is.finite(var_y), var_y > 0)

  ellipse_df <- plot_df %>%
    dplyr::semi_join(ellipse_groups, by = "ecological_preference_q")

  x_lab <- if (is.finite(explained[1])) sprintf("PCoA 1 (%.1f%%)", explained[1]) else "PCoA 1"
  y_lab <- if (is.finite(explained[2])) sprintf("PCoA 2 (%.1f%%)", explained[2]) else "PCoA 2"

  p <- ggplot(plot_df, aes(x = PCoA1, y = PCoA2, colour = ecological_preference_q)) +
    geom_point(size = 1.8, alpha = 0.85) +
    stat_ellipse(data = ellipse_df, linewidth = 0.3, alpha = 0.35, show.legend = FALSE) +
    scale_colour_manual(values = ECOLOGICAL_QUARTILE_COLORS, drop = FALSE) +
    labs(x = x_lab, y = y_lab, colour = "Ecological preference score") +
    fig_theme(base_size = 9, legend_position = "right")

  save_figure(p, filename, width = width, height = height, out_dir = SUPP_DIR)
  save_table(plot_df, paste0(filename, "_data.csv"))
  save_table(ellipse_groups, paste0(filename, "_ellipse_groups.csv"))
  p
}

run_maaslin2_for_table <- function(feature_mat, meta, cohort_key, level_label, stage_label) {
  if (!RUN_MAASLIN) return(tibble())
  if (!requireNamespace("Maaslin2", quietly = TRUE)) {
    stop("Package 'Maaslin2' is required when RUN_MAASLIN is TRUE.")
  }
  if (is.null(feature_mat)) return(tibble())

  aligned <- align_feature_to_meta(feature_mat, meta)
  ft <- t(aligned$feature)
  suppressWarnings(storage.mode(ft) <- "numeric")
  ft[!is.finite(ft)] <- 0
  ft <- ft[rowSums(ft, na.rm = TRUE) > 0, , drop = FALSE]
  ft <- ft[, colSums(ft, na.rm = TRUE) > 0, drop = FALSE]
  if (nrow(ft) < 5 || ncol(ft) < 2) return(tibble())

  design <- build_model_design(
    aligned$meta[rownames(ft), , drop = FALSE],
    cohort_key,
    stage_label,
    include_preferences = TRUE
  )
  cc <- stats::complete.cases(design)
  design <- drop_unused_factor_levels(design[cc, , drop = FALSE])
  ft <- ft[rownames(design), , drop = FALSE]
  if (nrow(ft) < 5 || ncol(ft) < 2 || ncol(design) < 1) return(tibble())

  if (cohort_key == "student" && stage_label == "stage2") {
    required_student_stage2 <- LIFESTYLE_CANDIDATES_STUDENT
    missing_stage2 <- setdiff(required_student_stage2, names(design))
    if (length(missing_stage2)) {
      stop(
        "Student cohort Stage 2 MaAsLin2 design does not include: ",
        paste(missing_stage2, collapse = ", "),
        call. = FALSE
      )
    }

    design$smoking <- droplevels(make_smoking_factor(design$smoking))
    design$clubmember <- droplevels(make_clubmember_factor(design$clubmember))
    design$alcohol_frequency <- droplevels(make_alcohol_frequency_factor(design$alcohol_frequency))

    low_level <- required_student_stage2[
      vapply(design[required_student_stage2], nlevels, integer(1)) < 2
    ]
    if (length(low_level)) {
      stop(
        "Student cohort Stage 2 covariate(s) have fewer than two observed levels after filtering: ",
        paste(low_level, collapse = ", "),
        call. = FALSE
      )
    }
  }

  write_design_audit(design, cohort_key, level_label, stage_label, phase = "final")

  cohort_dir <- file.path(RESULTS_DIR, cohort_config[[cohort_key]]$internal)
  maaslin_dir <- file.path(cohort_dir, paste0("maaslin_", stage_label, "_", tolower(level_label)))
  dir.create(cohort_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(maaslin_dir, recursive = TRUE, showWarnings = FALSE)

  readr::write_delim(
    meta_with_sample_id(design),
    file.path(cohort_dir, paste0("maaslin_design_", stage_label, "_", tolower(level_label), "_semicolon.csv")),
    delim = ";", na = ""
  )

  suppressWarnings(Maaslin2::Maaslin2(
    input_data = as.data.frame(ft, check.names = FALSE),
    input_metadata = as.data.frame(design, check.names = FALSE),
    output = maaslin_dir,
    normalization = "TSS",
    transform = "LOG",
    analysis_method = "LM",
    min_prevalence = 0.10,
    max_significance = FDR_THRESHOLD,
    standardize = FALSE,
    plot_heatmap = FALSE,
    plot_scatter = FALSE,
    cores = MAASLIN_CORES
  ))

  res_file <- file.path(maaslin_dir, "all_results.tsv")
  if (!file.exists(res_file)) return(tibble())

  out <- readr::read_tsv(res_file, show_col_types = FALSE) %>%
    janitor::clean_names() %>%
    dplyr::mutate(
      cohort = cohort_config[[cohort_key]]$cohort,
      cohort_key = cohort_key,
      level = tolower(level_label),
      stage = stage_label,
      n_samples = nrow(design),
      model_terms = paste(names(design), collapse = "; ")
    )

  readr::write_delim(
    out,
    file.path(cohort_dir, paste0("maaslin_", stage_label, "_", tolower(level_label), "_semicolon.csv")),
    delim = ";", na = ""
  )
  out
}
```

## 6.0 Alpha-diversity preference models

The descriptive between-cohort alpha-diversity figure above is complemented by within-cohort multivariable models. Shannon diversity and observed richness are modelled separately using the same Stage 1 and Stage 2 design logic as the community-composition and taxon-level analyses. Benjamini--Hochberg correction is applied across all preference coefficients for the two alpha-diversity outcomes within each cohort and stage.

```{r alpha-preference-models}
run_alpha_preference_models <- function(alpha_df, meta, cohort_key, stage_label) {
  if (is.null(alpha_df) || !nrow(alpha_df)) return(tibble())

  common <- intersect(alpha_df$sample_id, rownames(meta))
  if (length(common) < 5) return(tibble())

  meta2 <- meta[match(common, rownames(meta)), , drop = FALSE]
  rownames(meta2) <- common
  alpha2 <- alpha_df[match(common, alpha_df$sample_id), , drop = FALSE]

  design <- build_model_design(
    meta2,
    cohort_key = cohort_key,
    stage = stage_label,
    include_preferences = TRUE
  )
  alpha2 <- alpha2[match(rownames(design), alpha2$sample_id), , drop = FALSE]

  outcomes <- c("shannon", "richness")
  pref_terms <- preference_terms(meta2)

  purrr::map_dfr(outcomes, function(outcome_name) {
    if (!outcome_name %in% names(alpha2)) return(tibble())
    y <- numify(alpha2[[outcome_name]])
    dat <- dplyr::bind_cols(tibble(outcome_value = y), design)

    fit <- tryCatch(
      stats::lm(
        stats::reformulate(names(design), response = "outcome_value"),
        data = dat,
        na.action = stats::na.omit
      ),
      error = function(e) {
        warning(
          "Alpha-diversity model failed for ", cohort_key, "/", stage_label,
          "/", outcome_name, ": ", conditionMessage(e),
          call. = FALSE
        )
        NULL
      }
    )
    if (is.null(fit)) return(tibble())

    broom::tidy(fit, conf.int = TRUE) %>%
      dplyr::mutate(
        cohort = cohort_config[[cohort_key]]$cohort,
        cohort_key = cohort_key,
        stage = stage_label,
        outcome = dplyr::recode(
          outcome_name,
          shannon = "Shannon diversity",
          richness = "Observed richness"
        ),
        n_samples = stats::nobs(fit),
        model_terms = paste(names(design), collapse = "; "),
        is_preference = term %in% pref_terms
      )
  })
}

alpha_model_results_raw <- dplyr::bind_rows(
  run_alpha_preference_models(alpha_community, meta_community, "community", "stage1"),
  run_alpha_preference_models(alpha_community, meta_community, "community", "stage2"),
  run_alpha_preference_models(alpha_student, meta_student, "student", "stage1"),
  run_alpha_preference_models(alpha_student, meta_student, "student", "stage2")
)

if (!nrow(alpha_model_results_raw)) {
  stop("No alpha-diversity preference model could be fitted. Inspect alpha sample overlap and model-design audits.", call. = FALSE)
}

alpha_preference_adjusted <- alpha_model_results_raw %>%
  dplyr::filter(is_preference, is.finite(p.value)) %>%
  dplyr::group_by(cohort, stage) %>%
  dplyr::mutate(q_alpha = stats::p.adjust(p.value, method = "BH")) %>%
  dplyr::ungroup() %>%
  dplyr::select(cohort, stage, outcome, term, q_alpha)

alpha_model_results <- alpha_model_results_raw %>%
  dplyr::left_join(
    alpha_preference_adjusted,
    by = c("cohort", "stage", "outcome", "term")
  )

alpha_preference_results <- alpha_model_results %>%
  dplyr::filter(is_preference) %>%
  dplyr::mutate(
    preference = unname(PREFERENCE_LABELS[term]),
    preference = dplyr::if_else(is.na(preference), term, preference)
  )

alpha_preference_summary <- alpha_preference_results %>%
  dplyr::group_by(cohort, stage) %>%
  dplyr::summarise(
    n_tests = dplyr::n(),
    n_nominal = sum(p.value < NOMINAL_THRESHOLD, na.rm = TRUE),
    n_fdr = sum(q_alpha <= FDR_THRESHOLD, na.rm = TRUE),
    .groups = "drop"
  )

alpha_preference_fdr_hits <- alpha_preference_results %>%
  dplyr::filter(is.finite(q_alpha), q_alpha <= FDR_THRESHOLD) %>%
  dplyr::arrange(q_alpha, p.value)

save_table(alpha_model_results, "alpha_model_stage1_stage2_all.csv")
save_table(alpha_preference_results, "alpha_preference_associations.csv")
save_table(alpha_preference_summary, "alpha_preference_summary.csv")
save_table(alpha_preference_fdr_hits, "alpha_preference_fdr_hits.csv")
knitr::kable(alpha_preference_summary)
```

## 6.1 Ecological-preference PCoA plots

Colours represent cohort-specific quartiles of the standardized ecological-preference score, from `Q1 lowest` to `Q4 highest`. Missing scores are shown separately. Ellipses are visual guides and are drawn only for groups with sufficient non-degenerate data.

```{r pcoa-ecological-quartiles}
# Use genus-level profiles in both cohorts so the visual comparison is based on
# the same taxonomic resolution. Quartiles are calculated within each cohort.
p_pcoa_community_genus <- plot_pcoa_ecological(
  features$community_genus,
  meta_community,
  cohort_label = "Community cohort",
  level_label = "genus",
  filename = "pcoa_community_genus_ecological_quartiles"
)

p_pcoa_student_genus <- plot_pcoa_ecological(
  features$student_genus,
  meta_student,
  cohort_label = "Student cohort",
  level_label = "genus",
  filename = "pcoa_student_genus_ecological_quartiles"
)

p_pcoa_community_genus
p_pcoa_student_genus
```

## 6.2 PERMANOVA

```{r beta-permanova-stage1-stage2}
permanova_results <- dplyr::bind_rows(
  run_permanova(features$community_genus, meta_community, "community", "Genus", "stage1"),
  run_permanova(features$community_genus, meta_community, "community", "Genus", "stage2"),
  run_permanova(features$student_genus, meta_student, "student", "Genus", "stage1"),
  run_permanova(features$student_genus, meta_student, "student", "Genus", "stage2")
)

knitr::kable(permanova_results)
save_table(permanova_results, "permanova_genus_stage1_stage2.csv")
```

```{r permanova-preference-figure}
permanova_plot_df <- permanova_results %>%
  dplyr::filter(term %in% names(PREFERENCE_LABELS)) %>%
  dplyr::mutate(
    preference = unname(PREFERENCE_LABELS[term]),
    cohort = factor(cohort, levels = c("Community cohort", "Student cohort")),
    stage_label = factor(stage, levels = c("stage1", "stage2"), labels = c("Stage 1", "Stage 2")),
    nominal = .data[["Pr(>F)"]] < NOMINAL_THRESHOLD,
    partial_r2_percent = 100 * R2
  ) %>%
  dplyr::filter(is.finite(partial_r2_percent))

if (nrow(permanova_plot_df)) {
  p_permanova <- ggplot(
    permanova_plot_df,
    aes(
      x = partial_r2_percent,
      y = forcats::fct_reorder(preference, partial_r2_percent),
      colour = preference,
      shape = nominal
    )
  ) +
    geom_point(size = 2.3) +
    facet_grid(cohort ~ stage_label, scales = "free_y", drop = FALSE) +
    scale_colour_manual(values = PREFERENCE_COLORS, drop = FALSE) +
    scale_shape_manual(
      values = c(`FALSE` = 1, `TRUE` = 16),
      labels = c(`FALSE` = paste0("P >= ", NOMINAL_THRESHOLD), `TRUE` = paste0("P < ", NOMINAL_THRESHOLD))
    ) +
    labs(x = "Partial R-squared (%)", y = NULL, colour = "Preference", shape = "Nominal association") +
    fig_theme(base_size = 9)

  save_figure(p_permanova, "permanova_preferences_partial_r2_publication", width = 9.2, height = 6.0)
  save_table(permanova_plot_df, "permanova_preferences_partial_r2_publication_data.csv")
  p_permanova
}
```

## 6.3 MaAsLin2

```{r maaslin-stage1-stage2}
if (RUN_MAASLIN) {
  maaslin_result_list <- list(
    run_maaslin2_for_table(features$community_family, meta_community, "community", "family", "stage1"),
    run_maaslin2_for_table(features$community_family, meta_community, "community", "family", "stage2"),
    run_maaslin2_for_table(features$community_genus, meta_community, "community", "genus", "stage1"),
    run_maaslin2_for_table(features$community_genus, meta_community, "community", "genus", "stage2"),
    run_maaslin2_for_table(features$student_family, meta_student, "student", "family", "stage1"),
    run_maaslin2_for_table(features$student_family, meta_student, "student", "family", "stage2"),
    run_maaslin2_for_table(features$student_genus, meta_student, "student", "genus", "stage1"),
    run_maaslin2_for_table(features$student_genus, meta_student, "student", "genus", "stage2")
  )

  if (RUN_SPECIES_MAASLIN) {
    # Species-level analysis is scientifically supported for the Community-cohort
    # shotgun metagenomic profile only. Do not add Student-cohort 16S species calls.
    if (is.null(features$community_species)) {
      stop("Community-cohort shotgun species table is unavailable, but run_shotgun_species_maaslin is TRUE.", call. = FALSE)
    }
    maaslin_result_list <- c(
      maaslin_result_list,
      list(
        run_maaslin2_for_table(features$community_species, meta_community, "community", "species", "stage1"),
        run_maaslin2_for_table(features$community_species, meta_community, "community", "species", "stage2")
      )
    )
  }

  maaslin_results <- dplyr::bind_rows(maaslin_result_list)
} else {
  source_file <- if (!is.null(SOURCE_RUN_DIR)) {
    file.path(SOURCE_RUN_DIR, "tables", "maaslin_stage1_stage2_all.csv")
  } else NA_character_

  if (!is.na(source_file) && file.exists(source_file)) {
    maaslin_results <- readr::read_csv(source_file, show_col_types = FALSE) %>% janitor::clean_names()
    message("Loaded MaAsLin2 results from: ", source_file)
  } else {
    maaslin_results <- tibble()
    warning("MaAsLin2 was not run and no valid source_run_dir was supplied.", call. = FALSE)
  }
}

if (nrow(maaslin_results)) save_table(maaslin_results, "maaslin_stage1_stage2_all.csv")

maaslin_model_coverage <- if (nrow(maaslin_results)) {
  maaslin_results %>%
    dplyr::distinct(cohort, cohort_key, stage, level, n_samples, model_terms) %>%
    dplyr::arrange(cohort, stage, level)
} else {
  tibble(
    cohort = character(0), cohort_key = character(0), stage = character(0),
    level = character(0), n_samples = integer(0), model_terms = character(0)
  )
}
save_table(maaslin_model_coverage, "maaslin_model_coverage.csv")
knitr::kable(maaslin_model_coverage)

species_analysis_scope <- tibble(
  requested = RUN_SPECIES_MAASLIN,
  community_shotgun_species_models = if (nrow(maaslin_results)) sum(
    maaslin_results$cohort_key == "community" & tolower(maaslin_results$level) == "species",
    na.rm = TRUE
  ) else 0L,
  student_16s_species_models = if (nrow(maaslin_results)) sum(
    maaslin_results$cohort_key == "student" & tolower(maaslin_results$level) == "species",
    na.rm = TRUE
  ) else 0L,
  scope_valid = if (!RUN_SPECIES_MAASLIN) TRUE else (
    any(maaslin_model_coverage$cohort_key == "community" & tolower(maaslin_model_coverage$level) == "species") &&
      !any(maaslin_model_coverage$cohort_key == "student" & tolower(maaslin_model_coverage$level) == "species")
  )
)
save_table(species_analysis_scope, "species_analysis_scope.csv")
knitr::kable(species_analysis_scope)
```

```{r maaslin-model-audits}
student_stage2_covariate_design <- build_model_design(meta_student, "student", "stage2", include_preferences = FALSE)
student_stage2_covariate_design <- drop_unused_factor_levels(
  student_stage2_covariate_design[stats::complete.cases(student_stage2_covariate_design), , drop = FALSE]
)

student_stage2_full_design <- build_model_design(meta_student, "student", "stage2", include_preferences = TRUE)
student_stage2_full_design <- drop_unused_factor_levels(
  student_stage2_full_design[stats::complete.cases(student_stage2_full_design), , drop = FALSE]
)

student_stage2_required <- LIFESTYLE_CANDIDATES_STUDENT

student_stage2_check <- dplyr::bind_rows(
  purrr::map_dfr(student_stage2_required, function(v) {
    tibble(
      variable = v,
      present_covariate_design = v %in% names(student_stage2_covariate_design),
      factor_covariate_design = v %in% names(student_stage2_covariate_design) && is.factor(student_stage2_covariate_design[[v]]),
      n_levels_covariate_design = if (v %in% names(student_stage2_covariate_design)) nlevels(droplevels(student_stage2_covariate_design[[v]])) else 0L,
      present_full_design = v %in% names(student_stage2_full_design),
      factor_full_design = v %in% names(student_stage2_full_design) && is.factor(student_stage2_full_design[[v]]),
      n_levels_full_design = if (v %in% names(student_stage2_full_design)) nlevels(droplevels(student_stage2_full_design[[v]])) else 0L
    )
  }),
  tibble(
    variable = "__design_terms__",
    present_covariate_design = TRUE,
    factor_covariate_design = NA,
    n_levels_covariate_design = NA_integer_,
    present_full_design = TRUE,
    factor_full_design = NA,
    n_levels_full_design = NA_integer_
  )
)

student_stage2_design_terms <- tibble(
  design = c("covariates_only", "full_with_preferences"),
  terms = c(
    paste(names(student_stage2_covariate_design), collapse = "; "),
    paste(names(student_stage2_full_design), collapse = "; ")
  ),
  n_samples = c(nrow(student_stage2_covariate_design), nrow(student_stage2_full_design))
)

knitr::kable(student_stage2_check)
knitr::kable(student_stage2_design_terms)
save_table(student_stage2_check, "student_stage2_covariate_check.csv")
save_table(student_stage2_design_terms, "student_stage2_design_terms.csv")

stopifnot(
  all(student_stage2_check$present_covariate_design[student_stage2_check$variable %in% student_stage2_required]),
  all(student_stage2_check$factor_covariate_design[student_stage2_check$variable %in% student_stage2_required]),
  all(student_stage2_check$n_levels_covariate_design[student_stage2_check$variable %in% student_stage2_required] > 1),
  all(student_stage2_check$present_full_design[student_stage2_check$variable %in% student_stage2_required]),
  all(student_stage2_check$factor_full_design[student_stage2_check$variable %in% student_stage2_required]),
  all(student_stage2_check$n_levels_full_design[student_stage2_check$variable %in% student_stage2_required] > 1)
)

student_stage2_term_check <- if (nrow(maaslin_results)) {
  maaslin_results %>%
    dplyr::filter(cohort == "Student cohort", stage == "stage2") %>%
    dplyr::distinct(level, model_terms) %>%
    dplyr::mutate(
      smoking_in_model_terms = stringr::str_detect(model_terms, "(^|; )smoking($|; )"),
      clubmember_in_model_terms = stringr::str_detect(model_terms, "(^|; )clubmember($|; )"),
      alcohol_frequency_in_model_terms = stringr::str_detect(model_terms, "(^|; )alcohol_frequency($|; )"),
      all_student_stage2_covariates_present = smoking_in_model_terms & clubmember_in_model_terms & alcohol_frequency_in_model_terms
    )
} else {
  tibble(
    level = character(0), model_terms = character(0),
    smoking_in_model_terms = logical(0), clubmember_in_model_terms = logical(0),
    alcohol_frequency_in_model_terms = logical(0), all_student_stage2_covariates_present = logical(0)
  )
}

save_table(student_stage2_term_check, "student_stage2_model_term_check.csv")
if (nrow(student_stage2_term_check) && any(!student_stage2_term_check$all_student_stage2_covariates_present)) {
  stop("At least one Student cohort Stage 2 MaAsLin2 model is missing a required Stage 2 covariate.", call. = FALSE)
}

compare_student_stages <- function(df) {
  required <- c("cohort", "stage", "level", "feature", "metadata", "value", "coef")
  if (is.null(df) || !nrow(df) || !all(required %in% names(df))) return(tibble())
  df %>%
    dplyr::filter(cohort == "Student cohort", stage %in% c("stage1", "stage2")) %>%
    dplyr::select(level, feature, metadata, value, stage, coef) %>%
    tidyr::pivot_wider(names_from = stage, values_from = coef) %>%
    dplyr::filter(is.finite(stage1), is.finite(stage2)) %>%
    dplyr::group_by(level) %>%
    dplyr::summarise(
      n_common_coefficients = dplyr::n(),
      max_absolute_difference = max(abs(stage2 - stage1), na.rm = TRUE),
      median_absolute_difference = median(abs(stage2 - stage1), na.rm = TRUE),
      effectively_identical = max_absolute_difference <= 1e-12,
      .groups = "drop"
    )
}

student_stage_comparison <- compare_student_stages(maaslin_results)
knitr::kable(student_stage_comparison)
save_table(student_stage_comparison, "student_stage1_stage2_coefficient_comparison.csv")

if (nrow(student_stage_comparison) && any(student_stage_comparison$effectively_identical)) {
  stop("Student cohort Stage 1 and Stage 2 coefficients are exactly duplicated for at least one taxonomic level.", call. = FALSE)
}
```

# 7. Taxon-level figures

The main figure shows every taxon-level association passing the prespecified FDR threshold (`q <= 0.10`). Cohort and model stage are shown in a fixed 2-by-2 facet layout. Empty panels are labelled explicitly rather than assigned zero height; this prevents the non-finite viewport error that occurred when no Community-cohort association passed FDR correction. Preference colours are identical in the PERMANOVA, main taxon, and supplementary taxon figures.

```{r taxon-figure-functions}
pretty_stage <- function(x) {
  dplyr::recode(as.character(x), stage1 = "Stage 1", stage2 = "Stage 2", .default = as.character(x))
}

pretty_level <- function(x) stringr::str_to_title(as.character(x))

pretty_term <- function(metadata) {
  metadata <- as.character(metadata)
  out <- unname(PREFERENCE_LABELS[metadata])
  out[is.na(out) & metadata == "age"] <- "Age"
  out[is.na(out) & metadata == "bmi"] <- "BMI"
  out[is.na(out) & metadata == "gender"] <- "Gender"
  out[is.na(out) & metadata == "smoking"] <- "Smoking"
  out[is.na(out) & metadata == "alcohol"] <- "Alcohol"
  out[is.na(out) & metadata == "alcohol_frequency"] <- "Alcohol frequency"
  out[is.na(out) & metadata == "clubmember"] <- "Club membership"
  out[is.na(out)] <- "Other covariate"
  out
}

pretty_predictor_value <- function(metadata, value) {
  base <- pretty_term(metadata)
  value <- as.character(value)
  factor_term <- metadata %in% c("gender", "smoking", "marital_status", "clubmember", "alcohol_frequency")
  val_ok <- !is.na(value) & value != "" & value != metadata
  out <- base
  out[factor_term & val_ok] <- paste0(base[factor_term & val_ok], ": ", value[factor_term & val_ok])
  out
}

pretty_taxon_one <- function(one) {
  if (is.na(one) || !nzchar(stringr::str_trim(one))) return("Unassigned taxon")

  one <- as.character(one)
  one <- stringr::str_replace(one, "^(Bacteria|Archaea)[.;]", "")
  one <- stringr::str_replace_all(one, "_", " ")

  # mOTU-style labels often contain a readable organism name before the
  # technical ref/meta/ext_mOTU identifier. Preserve that readable part.
  if (stringr::str_detect(one, "(?i)(ref|meta|ext) mOTU")) {
    clean <- stringr::str_remove(one, "(?i)[.; ]+(ref|meta|ext) mOTU.*$")
    clean <- stringr::str_replace_all(clean, "[.;]+", " ")
    clean <- stringr::str_squish(clean)
    clean <- stringr::str_replace_all(clean, "\\bsp\\b", "sp.")
    clean <- stringr::str_replace_all(clean, "\\bgen\\b", "gen.")
    clean <- stringr::str_replace_all(clean, "\\bfam\\b", "fam.")

    # Prefer a full incertae-sedis label, then genus + sp./CAG, then a binomial.
    incertae <- stringr::str_extract(
      clean,
      "^[A-Za-z0-9-]+\\s+species\\s+incertae\\s+sedis"
    )
    if (!is.na(incertae)) return(stringr::str_trunc(incertae, width = 52))

    sp_label <- stringr::str_extract(
      clean,
      "^[A-Z][A-Za-z0-9-]+\\s+sp\\.(?:\\s+CAG\\s+[0-9]+)?"
    )
    if (!is.na(sp_label)) return(stringr::str_trunc(sp_label, width = 52))

    binomial <- stringr::str_extract(clean, "^[A-Z][A-Za-z0-9-]+\\s+[a-z][A-Za-z0-9-]+")
    if (!is.na(binomial) && !stringr::str_detect(binomial, "\\s+(gen|fam|species)$")) {
      return(stringr::str_trunc(binomial, width = 52))
    }

    # For ambiguous rank strings, use the last informative assignment.
    tokens <- unlist(stringr::str_split(clean, "\\s+"))
    tokens <- tokens[
      tokens != "" &
        !stringr::str_detect(tokens, "^(species|incertae|sedis|gen\\.|fam\\.|sp\\.|CAG|[0-9]+)$")
    ]
    if (length(tokens)) return(stringr::str_trunc(tail(tokens, 1), width = 52))
  }

  # GTDB / hierarchical labels: use the most specific informative rank.
  parts <- unlist(stringr::str_split(one, "[.;|]"))
  parts <- stringr::str_trim(parts)
  parts <- stringr::str_replace(parts, "^[dkpcofgs]__", "")
  parts <- parts[
    parts != "" &
      !stringr::str_detect(
        parts,
        "^(Bacteria|Archaea|incertae|sedis|species|gen|fam|order|class|phylum|unassigned|noHit|NA)$"
      )
  ]
  if (!length(parts)) return("Unassigned taxon")

  ans <- tail(parts, 1)
  ans <- stringr::str_squish(ans)
  stringr::str_trunc(ans, width = 52)
}

pretty_taxon <- function(x) {
  vapply(as.character(x), pretty_taxon_one, character(1))
}

prepare_maaslin_plot_data <- function(df) {
  if (is.null(df) || !nrow(df)) return(tibble())
  df <- df %>% janitor::clean_names()
  required <- c("feature", "metadata", "value", "coef", "stderr", "pval", "qval", "cohort", "level", "stage")
  missing <- setdiff(required, names(df))
  if (length(missing)) stop("MaAsLin2 results lack required columns: ", paste(missing, collapse = ", "))

  df %>%
    dplyr::transmute(
      cohort = factor(pretty_cohort(cohort), levels = c("Community cohort", "Student cohort")),
      stage = factor(pretty_stage(stage), levels = c("Stage 1", "Stage 2")),
      level = factor(pretty_level(level), levels = c("Family", "Genus", "Species")),
      feature = as.character(feature),
      taxon_short = pretty_taxon(feature),
      taxon_axis = paste0(taxon_short, " [", tolower(as.character(level)), "]"),
      metadata = as.character(metadata),
      value = as.character(value),
      estimate = as.numeric(coef),
      stderr = as.numeric(stderr),
      p = as.numeric(pval),
      q = as.numeric(qval),
      predictor = pretty_term(metadata),
      predictor_value = pretty_predictor_value(metadata, value),
      association_type = factor(ifelse(startsWith(metadata, "z_"), "Preference", "Covariate"), levels = c("Preference", "Covariate")),
      colour_key = predictor,
      conf_low = estimate - 1.96 * stderr,
      conf_high = estimate + 1.96 * stderr
    ) %>%
    dplyr::filter(
      !is.na(cohort), !is.na(stage), !is.na(level),
      is.finite(estimate), is.finite(stderr), stderr >= 0,
      is.finite(p), is.finite(conf_low), is.finite(conf_high)
    )
}

add_unique_axis_rows <- function(dat) {
  if (!nrow(dat)) return(dat)
  dat %>%
    dplyr::group_by(cohort, stage, taxon_axis) %>%
    dplyr::mutate(
      repeated_label = dplyr::n() > 1,
      axis_label = ifelse(repeated_label, paste0(taxon_axis, " -- ", predictor_value), taxon_axis)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::group_by(cohort, stage) %>%
    dplyr::arrange(estimate, .by_group = TRUE) %>%
    dplyr::mutate(panel_order = dplyr::row_number()) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      row_id = paste0("hit_", dplyr::row_number()),
      row_id = factor(row_id, levels = row_id)
    )
}

make_empty_taxon_panels <- function(dat, cohort_levels, stage_levels, message_text) {
  grid <- tidyr::expand_grid(
    cohort = factor(cohort_levels, levels = cohort_levels),
    stage = factor(stage_levels, levels = stage_levels)
  )
  occupied <- dat %>% dplyr::distinct(cohort, stage)
  empty <- dplyr::anti_join(grid, occupied, by = c("cohort", "stage"))
  if (!nrow(empty)) {
    return(tibble(
      cohort = factor(character(0), levels = cohort_levels),
      stage = factor(character(0), levels = stage_levels),
      row_id = factor(character(0)),
      estimate = numeric(0),
      axis_label = character(0),
      panel_message = character(0)
    ))
  }
  empty %>%
    dplyr::mutate(
      row_id = factor(
        paste0("empty_", as.integer(cohort), "_", as.integer(stage)),
        levels = paste0("empty_", as.integer(cohort), "_", as.integer(stage))
      ),
      estimate = 0,
      axis_label = "",
      panel_message = message_text
    )
}

colour_values_for_data <- function(keys) {
  values <- ASSOCIATION_COLORS
  missing_keys <- setdiff(unique(stats::na.omit(keys)), names(values))
  if (length(missing_keys)) {
    values <- c(values, stats::setNames(rep("#B3B3B3", length(missing_keys)), missing_keys))
  }
  values
}

# Diagnostic helper returning the exact data used in the main FDR figure.
get_fdr_taxa <- function(df, q_threshold = FDR_THRESHOLD) {
  prepared <- prepare_maaslin_plot_data(df)
  if (!nrow(prepared)) return(tibble())
  prepared %>%
    dplyr::filter(is.finite(q), q <= q_threshold) %>%
    add_unique_axis_rows()
}

plot_fdr_taxa <- function(df, filename = "taxa_fdr_significant_main") {
  prepared <- prepare_maaslin_plot_data(df)
  if (!nrow(prepared)) {
    message("No MaAsLin2 results available; main FDR figure not written.")
    return(NULL)
  }
  dat <- get_fdr_taxa(df, q_threshold = FDR_THRESHOLD)

  save_table(dat, paste0(filename, "_data.csv"))
  if (!nrow(dat)) {
    message(paste0("No taxon-level associations passed q <= ", FDR_THRESHOLD, "; main FDR figure not written."))
    return(NULL)
  }

  cohort_levels <- c("Community cohort", "Student cohort")
  stage_levels <- c("Stage 1", "Stage 2")
  empty <- make_empty_taxon_panels(dat, cohort_levels, stage_levels, paste0("No associations at q <= ", FDR_THRESHOLD))
  axis_labels <- c(
    stats::setNames(dat$axis_label, as.character(dat$row_id)),
    if (nrow(empty)) stats::setNames(empty$axis_label, as.character(empty$row_id)) else character(0)
  )

  q_dat <- dat %>%
    dplyr::mutate(
      q_label = paste0("q=", formatC(q, format = "f", digits = 3)),
      q_hjust = ifelse(estimate >= 0, -0.12, 1.12)
    )

  p <- ggplot() +
    geom_blank(data = empty, aes(x = estimate, y = row_id)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.35) +
    geom_errorbar(
      data = q_dat,
      aes(xmin = conf_low, xmax = conf_high, y = row_id, colour = colour_key),
      orientation = "y", width = 0.16, linewidth = 0.35
    ) +
    geom_point(
      data = q_dat,
      aes(x = estimate, y = row_id, colour = colour_key, shape = association_type),
      size = 2.4
    ) +
    geom_text(
      data = q_dat,
      aes(x = estimate, y = row_id, label = q_label, hjust = q_hjust),
      colour = "grey25", size = 2.5, show.legend = FALSE
    ) +
    geom_text(
      data = empty,
      aes(x = estimate, y = row_id, label = panel_message),
      colour = "grey45", size = 3, fontface = "italic"
    ) +
    facet_grid(cohort ~ stage, scales = "free_y", space = "fixed", drop = FALSE) +
    scale_y_discrete(labels = axis_labels) +
    scale_colour_manual(values = colour_values_for_data(q_dat$colour_key), drop = FALSE) +
    scale_shape_manual(values = c(Preference = 16, Covariate = 17), drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = c(0.12, 0.24))) +
    coord_cartesian(clip = "off") +
    labs(x = "MaAsLin2 coefficient", y = "Taxon", colour = "Association", shape = "Term type") +
    fig_theme(base_size = 9)

  panel_counts <- dat %>% dplyr::count(cohort, stage)
  max_rows <- max(c(panel_counts$n, 1L), na.rm = TRUE)
  height <- max(6.4, 2.8 + 2 * 0.38 * max_rows)

  panel_status <- tidyr::expand_grid(cohort = cohort_levels, stage = stage_levels) %>%
    dplyr::left_join(panel_counts, by = c("cohort", "stage")) %>%
    dplyr::mutate(n = tidyr::replace_na(n, 0L))
  save_table(panel_status, paste0(filename, "_panel_status.csv"))
  save_figure(p, filename, width = 10.2, height = height)
  p
}

plot_nominal_taxa <- function(df, level_to_plot, filename, n_per_panel = TOP_NOMINAL_PER_PANEL) {
  prepared <- prepare_maaslin_plot_data(df)
  if (!nrow(prepared)) {
    message("No MaAsLin2 results available; nominal ", level_to_plot, " figure not written.")
    return(NULL)
  }
  all_level <- prepared %>%
    dplyr::filter(as.character(level) == stringr::str_to_title(level_to_plot))

  if (!nrow(all_level)) {
    message("No ", level_to_plot, " results are available.")
    return(NULL)
  }

  dat_all <- all_level %>%
    dplyr::filter(association_type == "Preference", p < NOMINAL_THRESHOLD)
  save_table(dat_all, paste0(filename, "_all_nominal.csv"))

  if (!nrow(dat_all)) {
    message("No nominal preference associations available for ", level_to_plot, ".")
    return(NULL)
  }

  dat <- dat_all %>%
    dplyr::group_by(cohort, stage) %>%
    dplyr::arrange(p, dplyr::desc(abs(estimate)), .by_group = TRUE) %>%
    dplyr::slice_head(n = n_per_panel) %>%
    dplyr::ungroup() %>%
    add_unique_axis_rows()

  save_table(dat, paste0(filename, "_data.csv"))

  cohort_levels <- levels(droplevels(all_level$cohort))
  if (!length(cohort_levels)) cohort_levels <- unique(as.character(dat$cohort))
  stage_levels <- c("Stage 1", "Stage 2")
  empty <- make_empty_taxon_panels(dat, cohort_levels, stage_levels, "No nominal preference associations")
  axis_labels <- c(
    stats::setNames(dat$axis_label, as.character(dat$row_id)),
    if (nrow(empty)) stats::setNames(empty$axis_label, as.character(empty$row_id)) else character(0)
  )

  p <- ggplot() +
    geom_blank(data = empty, aes(x = estimate, y = row_id)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.35) +
    geom_point(
      data = dat,
      aes(x = estimate, y = row_id, colour = predictor),
      size = 2.2, alpha = 0.95
    ) +
    geom_text(
      data = empty,
      aes(x = estimate, y = row_id, label = panel_message),
      colour = "grey45", size = 3, fontface = "italic"
    ) +
    facet_grid(cohort ~ stage, scales = "free_y", space = "fixed", drop = FALSE) +
    scale_y_discrete(labels = axis_labels) +
    scale_colour_manual(values = PREFERENCE_COLORS, drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = c(0.10, 0.12))) +
    labs(x = "MaAsLin2 coefficient", y = "Taxon", colour = "Preference") +
    fig_theme(base_size = 8)

  panel_counts <- dat %>% dplyr::count(cohort, stage)
  max_rows <- max(c(panel_counts$n, 1L), na.rm = TRUE)
  n_cohort_rows <- max(length(cohort_levels), 1L)
  height <- max(5.4, 2.6 + n_cohort_rows * 0.29 * max_rows)

  save_figure(p, filename, width = 10.0, height = height, out_dir = SUPP_DIR)
  p
}
```

```{r taxon-fdr-diagnostics}
taxa_fdr <- get_fdr_taxa(maaslin_results)
taxon_fdr_diagnostics <- if (nrow(taxa_fdr)) {
  taxa_fdr %>%
    dplyr::count(cohort, stage, level, association_type, name = "n_associations")
} else {
  tibble(
    cohort = character(0), stage = character(0), level = character(0),
    association_type = character(0), n_associations = integer(0)
  )
}
knitr::kable(taxon_fdr_diagnostics)
save_table(taxon_fdr_diagnostics, "taxa_fdr_significant_main_diagnostics.csv")
```

```{r taxon-main-fdr-figure}
p_taxa_fdr <- plot_fdr_taxa(maaslin_results, "taxa_fdr_significant_main")
p_taxa_fdr
```

```{r taxon-exploratory-supplement}
p_taxa_nominal_family <- plot_nominal_taxa(
  maaslin_results, "Family", "taxa_nominal_family_supplement"
)
p_taxa_nominal_genus <- plot_nominal_taxa(
  maaslin_results, "Genus", "taxa_nominal_genus_supplement"
)
p_taxa_nominal_species <- if (RUN_SPECIES_MAASLIN || any(tolower(maaslin_results$level) == "species")) {
  plot_nominal_taxa(maaslin_results, "Species", "taxa_nominal_species_supplement")
} else NULL

p_taxa_nominal_family
p_taxa_nominal_genus
p_taxa_nominal_species
```

```{r taxon-association-summary}
taxon_plot_data_all <- prepare_maaslin_plot_data(maaslin_results)

if (nrow(taxon_plot_data_all)) {
  taxon_association_summary <- taxon_plot_data_all %>%
    dplyr::summarise(
      n_results = dplyr::n(),
      n_fdr_total = sum(q <= FDR_THRESHOLD, na.rm = TRUE),
      n_fdr_preferences = sum(q <= FDR_THRESHOLD & association_type == "Preference", na.rm = TRUE),
      n_fdr_covariates = sum(q <= FDR_THRESHOLD & association_type == "Covariate", na.rm = TRUE),
      n_nominal_preferences = sum(p < NOMINAL_THRESHOLD & association_type == "Preference", na.rm = TRUE)
    )
} else {
  taxon_association_summary <- tibble(
    n_results = 0L,
    n_fdr_total = 0L,
    n_fdr_preferences = 0L,
    n_fdr_covariates = 0L,
    n_nominal_preferences = 0L
  )
}

knitr::kable(taxon_association_summary)
save_table(taxon_association_summary, "taxon_association_summary.csv")
```

# 8. Community-cohort metabolomics

Metabolite models use the same Stage 2 design logic as the taxon-level models. Benjamini--Hochberg correction is applied across metabolites separately for each model term. The forest plot displays standardized-outcome coefficients for the strongest self-control associations and retains 3-IAA if it is present, without hard-coding its significance. Raw-scale coefficients remain available in the exported table.

```{r metabolomics-functions}
load_metabolome <- function() {
  path <- file.path(DATA_DIR, cohort_config$community$metabolome_file)
  if (!file.exists(path)) return(NULL)
  e <- new.env(parent = emptyenv())
  load(path, envir = e)
  if (!exists("mtblo", envir = e, inherits = FALSE)) {
    stop("Expected object 'mtblo' in ", path)
  }
  as.data.frame(get("mtblo", envir = e), check.names = FALSE)
}

align_metabolome_to_meta <- function(mtb, meta) {
  mtb <- as.data.frame(mtb, check.names = FALSE)
  meta_ids <- coerce_id(rownames(meta))
  row_ids <- coerce_id(rownames(mtb))
  col_ids <- coerce_id(colnames(mtb))
  row_common <- intersect(meta_ids, row_ids)
  col_common <- intersect(meta_ids, col_ids)

  if (length(row_common) >= length(col_common) && length(row_common) > 0) {
    common <- row_common
    mtb2 <- mtb[match(common, row_ids), , drop = FALSE]
  } else if (length(col_common) > 0) {
    mtb_t <- as.data.frame(t(as.matrix(mtb)), check.names = FALSE)
    row_ids_t <- coerce_id(rownames(mtb_t))
    common <- intersect(meta_ids, row_ids_t)
    mtb2 <- mtb_t[match(common, row_ids_t), , drop = FALSE]
  } else {
    stop("No shared sample IDs between metabolomics and Community-cohort metadata.")
  }

  meta2 <- meta[match(common, meta_ids), , drop = FALSE]
  rownames(meta2) <- common
  rownames(mtb2) <- common
  names(mtb2) <- make.unique(names(mtb2))
  list(metabolites = mtb2, meta = meta2, common = common)
}

run_metabolomics_analysis <- function(mtb, meta) {
  if (is.null(mtb)) return(tibble())
  aligned <- align_metabolome_to_meta(mtb, meta)
  mtb2 <- aligned$metabolites
  mtb2[] <- lapply(mtb2, numify)

  keep_metabolites <- vapply(mtb2, function(x) {
    y <- as.numeric(x)
    sum(is.finite(y)) >= 5 && stats::sd(y, na.rm = TRUE) > 0
  }, logical(1))
  mtb2 <- mtb2[, keep_metabolites, drop = FALSE]

  design <- build_model_design(aligned$meta, "community", "stage2", include_preferences = TRUE)
  cc <- stats::complete.cases(design)
  design <- drop_unused_factor_levels(design[cc, , drop = FALSE])
  mtb2 <- mtb2[rownames(design), , drop = FALSE]
  write_design_audit(design, "community", "metabolomics", "stage2", phase = "final")

  if (nrow(design) < 5 || ncol(design) < 1 || ncol(mtb2) < 1) return(tibble())

  result <- purrr::map_dfr(names(mtb2), function(metabolite_name) {
    y <- numify(mtb2[[metabolite_name]])
    y_sd <- stats::sd(y, na.rm = TRUE)
    if (sum(is.finite(y)) < 5 || !is.finite(y_sd) || y_sd == 0) return(tibble())
    dat <- dplyr::bind_cols(tibble(y = y), design)
    fit <- tryCatch(
      stats::lm(stats::reformulate(names(design), response = "y"), data = dat, na.action = stats::na.omit),
      error = function(e) NULL
    )
    if (is.null(fit)) return(tibble())
    broom::tidy(fit, conf.int = TRUE) %>%
      dplyr::mutate(
        metabolite = metabolite_name,
        cohort = "Community cohort",
        stage = "stage2",
        n_samples = stats::nobs(fit),
        model_terms = paste(names(design), collapse = "; "),
        outcome_sd = y_sd,
        estimate_std = dplyr::if_else(term == "(Intercept)", NA_real_, estimate / y_sd),
        std_error_std = dplyr::if_else(term == "(Intercept)", NA_real_, std.error / y_sd),
        conf_low_std = dplyr::if_else(term == "(Intercept)", NA_real_, conf.low / y_sd),
        conf_high_std = dplyr::if_else(term == "(Intercept)", NA_real_, conf.high / y_sd)
      )
  })

  if (nrow(result)) {
    result <- result %>%
      dplyr::group_by(term) %>%
      dplyr::mutate(q_term = stats::p.adjust(p.value, method = "BH")) %>%
      dplyr::ungroup()
  }
  result
}

pretty_metabolite <- function(x) {
  x <- as.character(x)
  x <- stringr::str_replace_all(x, "^X3_IAA$|^3_IAA$|^IAA$", "3-IAA")
  x <- stringr::str_replace_all(x, "^t4_OH_Pro$", "trans-4-hydroxyproline")
  x <- stringr::str_replace_all(x, "Cer_d18_1_20_0", "Cer(d18:1/20:0)")
  x <- stringr::str_replace_all(x, "DG_18_1_20_3", "DG(18:1/20:3)")
  x <- stringr::str_replace_all(x, "_", ":")
  x
}

plot_metabolite_selfcontrol <- function(results, filename = "metabolite_selfcontrol_forest_publication", top_n = 10) {
  if (is.null(results) || !nrow(results)) return(NULL)
  dat <- results %>%
    janitor::clean_names() %>%
    dplyr::filter(term == "z_selfcontrol") %>%
    dplyr::mutate(
      is_iaa = stringr::str_detect(stringr::str_to_lower(metabolite), "x3_iaa|3_iaa|3-iaa|indole.?3.?acetic|^iaa$"),
      metabolite_label = pretty_metabolite(metabolite),
      rank_metric = dplyr::coalesce(q_term, p_value),
      highlight = ifelse(is_iaa, "3-IAA", "Other metabolites")
    ) %>%
    dplyr::filter(
      is.finite(estimate_std), is.finite(conf_low_std), is.finite(conf_high_std),
      is.finite(p_value)
    )

  if (!nrow(dat)) return(NULL)
  top <- dat %>%
    dplyr::arrange(rank_metric, p_value, dplyr::desc(abs(estimate_std))) %>%
    dplyr::slice_head(n = top_n)
  if (any(dat$is_iaa) && !any(top$is_iaa)) {
    top <- dplyr::bind_rows(top, dat %>% dplyr::filter(is_iaa) %>% dplyr::arrange(rank_metric, p_value) %>% dplyr::slice_head(n = 1))
  }
  top <- top %>%
    dplyr::distinct(metabolite, .keep_all = TRUE) %>%
    dplyr::mutate(metabolite_f = forcats::fct_reorder(metabolite_label, estimate_std))

  p <- ggplot(top, aes(x = estimate_std, y = metabolite_f, colour = highlight)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.35) +
    geom_errorbar(
      aes(xmin = conf_low_std, xmax = conf_high_std),
      orientation = "y", width = 0.16, linewidth = 0.35
    ) +
    geom_point(size = 2.2) +
    scale_colour_manual(values = METABOLITE_COLORS, drop = FALSE) +
    labs(x = "Coefficient per metabolite SD", y = "Metabolite", colour = NULL) +
    fig_theme(base_size = 9)

  save_figure(p, filename, width = 6.8, height = 4.8)
  save_table(top %>% dplyr::select(-metabolite_f), paste0(filename, "_data.csv"))
  p
}
```

```{r metabolomics-analysis}
if (RUN_METABOLOMICS) {
  metabolomics_results <- run_metabolomics_analysis(load_metabolome(), meta_community)
} else {
  source_file <- if (!is.null(SOURCE_RUN_DIR)) file.path(SOURCE_RUN_DIR, "tables", "metabolomics_all.csv") else NA_character_
  if (!is.na(source_file) && file.exists(source_file)) {
    metabolomics_results <- readr::read_csv(source_file, show_col_types = FALSE)
  } else {
    metabolomics_results <- tibble()
    warning("Metabolomics was not run and no valid source result was supplied.", call. = FALSE)
  }
}

if (nrow(metabolomics_results)) {
  save_table(metabolomics_results, "metabolomics_all.csv")
  metabolomics_dir <- file.path(RESULTS_DIR, "hamburg")
  dir.create(metabolomics_dir, recursive = TRUE, showWarnings = FALSE)
  readr::write_delim(
    metabolomics_results,
    file.path(metabolomics_dir, "metabolomics_semicolon.csv"),
    delim = ";", na = ""
  )
}

metabolomics_fdr_hits <- if (nrow(metabolomics_results)) {
  metabolomics_results %>%
    dplyr::filter(startsWith(term, "z_"), is.finite(q_term), q_term <= FDR_THRESHOLD) %>%
    dplyr::arrange(q_term, p.value)
} else {
  tibble()
}

metabolomics_summary <- tibble(
  n_metabolites = if (nrow(metabolomics_results)) dplyr::n_distinct(metabolomics_results$metabolite) else 0L,
  n_model_rows = nrow(metabolomics_results),
  n_preference_fdr_hits = nrow(metabolomics_fdr_hits),
  iaa_present = if (nrow(metabolomics_results)) any(
    stringr::str_detect(
      stringr::str_to_lower(metabolomics_results$metabolite),
      "x3_iaa|3_iaa|3-iaa|indole.?3.?acetic|^iaa$"
    )
  ) else FALSE
)

save_table(metabolomics_fdr_hits, "metabolomics_preference_fdr_hits.csv")
save_table(metabolomics_summary, "metabolomics_summary.csv")
knitr::kable(metabolomics_fdr_hits)
knitr::kable(metabolomics_summary)
```

```{r metabolomics-figure}
p_metabolite_selfcontrol <- plot_metabolite_selfcontrol(metabolomics_results)
p_metabolite_selfcontrol
```

# 9. Harmonized Bacillota/Bacteroidota ratio

The Bacillota/Bacteroidota ratio is retained as a secondary exploratory analysis. It is calculated independently in each cohort from the corresponding rarefied **phylum-level** abundance table. Historical phylum names are harmonized (`Firmicutes` to `Bacillota`; `Bacteroidetes` to `Bacteroidota`). The inferential outcome is the natural logarithm of the ratio after adding a transparent 0.5-count pseudocount to both phyla; the untransformed ratio is also exported for descriptive use. The same Stage 1 and Stage 2 covariate structures are used as in the other within-cohort models. A null result is retained and reported rather than used as a reason to omit the analysis.

```{r fb-ratio-functions}
FB_PSEUDOCOUNT <- 0.5

phylum_token_match <- function(labels, token) {
  labels <- stringr::str_to_lower(as.character(labels))
  labels <- stringr::str_replace_all(labels, "firmicutes", "bacillota")
  labels <- stringr::str_replace_all(labels, "bacteroidetes", "bacteroidota")
  stringr::str_detect(labels, paste0("(^|[^a-z])", token, "([^a-z]|$)"))
}

prepare_fb_ratio <- function(phylum_mat, meta, cohort_key, source_file) {
  if (is.null(phylum_mat) || !nrow(phylum_mat) || !ncol(phylum_mat)) {
    stop("Phylum table is empty for ", cohort_key, ".", call. = FALSE)
  }

  taxa <- rownames(phylum_mat)
  bacillota_ix <- phylum_token_match(taxa, "bacillota")
  bacteroidota_ix <- phylum_token_match(taxa, "bacteroidota")

  if (!any(bacillota_ix) || !any(bacteroidota_ix)) {
    stop(
      "Could not identify both Bacillota and Bacteroidota in the phylum table for ",
      cohort_key, ". Bacillota rows: ", sum(bacillota_ix),
      "; Bacteroidota rows: ", sum(bacteroidota_ix),
      call. = FALSE
    )
  }

  aligned <- align_feature_to_meta(phylum_mat, meta)
  mat <- aligned$feature

  bacillota_count <- colSums(mat[bacillota_ix, , drop = FALSE], na.rm = TRUE)
  bacteroidota_count <- colSums(mat[bacteroidota_ix, , drop = FALSE], na.rm = TRUE)
  total_count <- colSums(mat, na.rm = TRUE)

  data <- tibble(
    sample_id = aligned$common,
    cohort = cohort_config[[cohort_key]]$cohort,
    cohort_key = cohort_key,
    bacillota_count = as.numeric(bacillota_count),
    bacteroidota_count = as.numeric(bacteroidota_count),
    total_phylum_count = as.numeric(total_count),
    bacillota_relative = dplyr::if_else(total_count > 0, bacillota_count / total_count, NA_real_),
    bacteroidota_relative = dplyr::if_else(total_count > 0, bacteroidota_count / total_count, NA_real_),
    fb_ratio_raw = dplyr::if_else(bacteroidota_count > 0, bacillota_count / bacteroidota_count, NA_real_),
    fb_ratio_pseudocount = (bacillota_count + FB_PSEUDOCOUNT) / (bacteroidota_count + FB_PSEUDOCOUNT),
    log_fb_ratio = log(fb_ratio_pseudocount),
    source_file = source_file
  )

  diagnostics <- tibble(
    cohort = cohort_config[[cohort_key]]$cohort,
    cohort_key = cohort_key,
    source_file = source_file,
    n_samples = nrow(data),
    n_bacillota_rows = sum(bacillota_ix),
    n_bacteroidota_rows = sum(bacteroidota_ix),
    bacillota_labels = paste(taxa[bacillota_ix], collapse = " | "),
    bacteroidota_labels = paste(taxa[bacteroidota_ix], collapse = " | "),
    n_zero_bacillota = sum(data$bacillota_count == 0, na.rm = TRUE),
    n_zero_bacteroidota = sum(data$bacteroidota_count == 0, na.rm = TRUE),
    n_finite_log_ratio = sum(is.finite(data$log_fb_ratio))
  )

  list(data = data, diagnostics = diagnostics)
}

run_fb_preference_models <- function(fb_df, meta, cohort_key, stage_label) {
  if (is.null(fb_df) || !nrow(fb_df)) return(tibble())

  common <- intersect(fb_df$sample_id, rownames(meta))
  if (length(common) < 5) return(tibble())

  meta2 <- meta[match(common, rownames(meta)), , drop = FALSE]
  rownames(meta2) <- common
  fb2 <- fb_df[match(common, fb_df$sample_id), , drop = FALSE]

  design <- build_model_design(
    meta2,
    cohort_key = cohort_key,
    stage = stage_label,
    include_preferences = TRUE
  )
  fb2 <- fb2[match(rownames(design), fb2$sample_id), , drop = FALSE]

  dat <- dplyr::bind_cols(
    tibble(outcome_value = numify(fb2$log_fb_ratio)),
    design
  )
  pref_terms <- preference_terms(meta2)

  fit <- tryCatch(
    stats::lm(
      stats::reformulate(names(design), response = "outcome_value"),
      data = dat,
      na.action = stats::na.omit
    ),
    error = function(e) {
      warning(
        "Bacillota/Bacteroidota model failed for ", cohort_key, "/", stage_label,
        ": ", conditionMessage(e), call. = FALSE
      )
      NULL
    }
  )
  if (is.null(fit)) return(tibble())

  broom::tidy(fit, conf.int = TRUE) %>%
    dplyr::mutate(
      cohort = cohort_config[[cohort_key]]$cohort,
      cohort_key = cohort_key,
      stage = stage_label,
      outcome = "log(Bacillota/Bacteroidota)",
      n_samples = stats::nobs(fit),
      pseudocount = FB_PSEUDOCOUNT,
      model_terms = paste(names(design), collapse = "; "),
      is_preference = term %in% pref_terms
    )
}
```

```{r fb-ratio-analysis}
if (RUN_FB_RATIO) {

  fb_community_obj <- prepare_fb_ratio(
    features$community_phylum,
    meta_community,
    "community",
    cohort_config$community$rarefied_tables$phylum
  )

  fb_student_obj <- prepare_fb_ratio(
    features$student_phylum,
    meta_student,
    "student",
    cohort_config$student$rarefied_tables$phylum
  )

  fb_ratio_data <- dplyr::bind_rows(
    fb_community_obj$data,
    fb_student_obj$data
  )

  fb_ratio_diagnostics <- dplyr::bind_rows(
    fb_community_obj$diagnostics,
    fb_student_obj$diagnostics
  )

  fb_model_results_raw <- dplyr::bind_rows(
    run_fb_preference_models(
      fb_community_obj$data,
      meta_community,
      "community",
      "stage1"
    ),
    run_fb_preference_models(
      fb_community_obj$data,
      meta_community,
      "community",
      "stage2"
    ),
    run_fb_preference_models(
      fb_student_obj$data,
      meta_student,
      "student",
      "stage1"
    ),
    run_fb_preference_models(
      fb_student_obj$data,
      meta_student,
      "student",
      "stage2"
    )
  )

  fb_preference_adjusted <- fb_model_results_raw %>%
    dplyr::filter(is_preference, is.finite(p.value)) %>%
    dplyr::group_by(cohort, stage) %>%
    dplyr::mutate(
      q_fb = stats::p.adjust(p.value, method = "BH")
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(cohort, stage, term, q_fb)

  fb_model_results <- fb_model_results_raw %>%
    dplyr::left_join(
      fb_preference_adjusted,
      by = c("cohort", "stage", "term")
    )

  fb_preference_results <- fb_model_results %>%
    dplyr::filter(is_preference) %>%
    dplyr::mutate(
      preference = unname(PREFERENCE_LABELS[term]),
      preference = dplyr::if_else(
        is.na(preference),
        term,
        preference
      )
    )

  fb_preference_summary <- fb_preference_results %>%
    dplyr::group_by(cohort, stage) %>%
    dplyr::summarise(
      n_tests = dplyr::n(),
      n_nominal = sum(
        p.value < NOMINAL_THRESHOLD,
        na.rm = TRUE
      ),
      n_fdr = sum(
        q_fb <= FDR_THRESHOLD,
        na.rm = TRUE
      ),
      .groups = "drop"
    )

  fb_preference_fdr_hits <- fb_preference_results %>%
    dplyr::filter(
      is.finite(q_fb),
      q_fb <= FDR_THRESHOLD
    ) %>%
    dplyr::arrange(q_fb, p.value)

  save_table(
    fb_ratio_data,
    "fb_ratio_harmonized_sample_data.csv"
  )

  save_table(
    fb_ratio_diagnostics,
    "fb_ratio_harmonized_diagnostics.csv"
  )

  save_table(
    fb_model_results,
    "fb_ratio_stage1_stage2_all.csv"
  )

  save_table(
    fb_preference_results,
    "fb_ratio_preference_associations.csv"
  )

  save_table(
    fb_preference_summary,
    "fb_ratio_preference_summary.csv"
  )

  save_table(
    fb_preference_fdr_hits,
    "fb_ratio_preference_fdr_hits.csv"
  )

  fb_plot_df <- fb_preference_results %>%
    dplyr::filter(
      is.finite(estimate),
      is.finite(conf.low),
      is.finite(conf.high)
    ) %>%
    dplyr::mutate(
      cohort = factor(
        cohort,
        levels = c(
          "Community cohort",
          "Student cohort"
        )
      ),
      stage_label = factor(
        stage,
        levels = c("stage1", "stage2"),
        labels = c("Stage 1", "Stage 2")
      ),
      preference = factor(
        preference,
        levels = names(PREFERENCE_COLORS)
      ),
      nominal = p.value < NOMINAL_THRESHOLD
    )

  if (nrow(fb_plot_df) > 0) {

    p_fb_ratio_preferences <- ggplot(
      fb_plot_df,
      aes(
        x = estimate,
        y = forcats::fct_reorder(preference, estimate),
        colour = preference,
        shape = nominal
      )
    ) +
      geom_vline(
        xintercept = 0,
        linetype = 2,
        colour = "grey60"
      ) +
      geom_errorbar(
        aes(
          xmin = conf.low,
          xmax = conf.high
        ),
        orientation = "y",
        width = 0.16
      ) +
      geom_point(size = 2.0) +
      facet_grid(
        cohort ~ stage_label,
        scales = "free_y",
        drop = FALSE
      ) +
      scale_colour_manual(
        values = PREFERENCE_COLORS,
        drop = FALSE
      ) +
      scale_shape_manual(
        values = c(
          `FALSE` = 1,
          `TRUE` = 16
        ),
        labels = c(
          `FALSE` = paste0(
            "P >= ",
            NOMINAL_THRESHOLD
          ),
          `TRUE` = paste0(
            "P < ",
            NOMINAL_THRESHOLD
          )
        )
      ) +
      labs(
        x = "Coefficient for log Bacillota/Bacteroidota ratio",
        y = NULL,
        colour = "Preference",
        shape = "Nominal association"
      ) +
      fig_theme(base_size = 8.5)

    save_figure(
      p_fb_ratio_preferences,
      "fb_ratio_preference_associations_supplement",
      width = 9.4,
      height = 7.2,
      out_dir = SUPP_DIR
    )

    print(p_fb_ratio_preferences)

  } else {

    p_fb_ratio_preferences <- NULL
    message(
      "No finite Bacillota/Bacteroidota preference estimates available for plotting."
    )
  }

  knitr::kable(fb_ratio_diagnostics)
  knitr::kable(fb_preference_summary)
  knitr::kable(fb_preference_fdr_hits)

} else {

  fb_ratio_data <- tibble::tibble()
  fb_ratio_diagnostics <- tibble::tibble()
  fb_model_results <- tibble::tibble()
  fb_preference_results <- tibble::tibble()
  fb_preference_summary <- tibble::tibble()
  fb_preference_fdr_hits <- tibble::tibble()
  p_fb_ratio_preferences <- NULL

  message(
    "Bacillota/Bacteroidota ratio analysis disabled by run_fb_ratio: false"
  )
}
```

# 10. Manuscript numbers and running results summary

```{r manuscript-numbers}
manuscript_numbers <- tibble(
  quantity = c(
    "n_community_metadata",
    "n_student_metadata",
    "n_community_preference_id_overlap",
    "n_community_alpha",
    "n_student_alpha",
    "n_alpha_preference_fdr_hits",
    "n_taxon_fdr_hits_total",
    "n_taxon_fdr_preference_hits",
    "n_fb_ratio_preference_fdr_hits",
    "n_metabolite_preference_fdr_hits"
  ),
  value = c(
    nrow(meta_community),
    nrow(meta_student),
    community_preference_linkage_audit$n[
      community_preference_linkage_audit$dataset == "IDs shared by meta_dummy and preference file"
    ],
    sum(alpha_main$cohort == "Community cohort"),
    sum(alpha_main$cohort == "Student cohort"),
    nrow(alpha_preference_fdr_hits),
    if (nrow(taxon_plot_data_all)) sum(taxon_plot_data_all$q <= FDR_THRESHOLD, na.rm = TRUE) else 0L,
    if (nrow(taxon_plot_data_all)) sum(taxon_plot_data_all$q <= FDR_THRESHOLD & taxon_plot_data_all$association_type == "Preference", na.rm = TRUE) else 0L,
    nrow(fb_preference_fdr_hits),
    nrow(metabolomics_fdr_hits)
  )
)
knitr::kable(manuscript_numbers)
save_table(manuscript_numbers, "manuscript_numbers.csv")
```

```{r manuscript-summary, results='asis'}
summary_lines <- c(
  "## Manuscript-ready running summary",
  "",
  paste0("The Community cohort contains ", nrow(meta_community), " harmonized metadata records."),
  "",
  paste0("The Student cohort contains ", nrow(meta_student), " harmonized metadata records."),
  "",
  paste0("Alpha diversity is available for ", sum(alpha_main$cohort == "Community cohort"), " Community-cohort and ", sum(alpha_main$cohort == "Student cohort"), " Student-cohort samples."),
  "",
  paste0("The within-cohort alpha-diversity models contain ", nrow(alpha_preference_fdr_hits), " preference associations at q <= ", FDR_THRESHOLD, "."),
  "",
  paste0("The current MaAsLin2 output contains ", manuscript_numbers$value[manuscript_numbers$quantity == "n_taxon_fdr_hits_total"], " taxon-level associations at q <= ", FDR_THRESHOLD, ", including ", manuscript_numbers$value[manuscript_numbers$quantity == "n_taxon_fdr_preference_hits"], " preference-related associations."),
  "",
  paste0("The harmonized Bacillota/Bacteroidota analysis contains ", nrow(fb_preference_fdr_hits), " preference associations at q <= ", FDR_THRESHOLD, "."),
  "",
  paste0("The current metabolomics output contains ", nrow(metabolomics_fdr_hits), " preference-related metabolite associations at q <= ", FDR_THRESHOLD, "."),
  "",
  "Student cohort Stage 2 models include smoking, club membership, and alcohol-consumption frequency as categorical covariates and are checked against exact duplication of Stage 1 coefficients.",
  "",
  "Species-level MaAsLin2 is restricted to the Community-cohort shotgun data; Student-cohort 16S profiles are analysed at family and genus level only.",
  "",
  "The harmonized Bacillota/Bacteroidota ratio is retained as a secondary exploratory analysis, including when the result is null."
)
cat(paste(summary_lines, collapse = "\n"))
writeLines(summary_lines, file.path(TABLES_DIR, "manuscript_running_summary.md"))
```

# 11. Output validation and session information

```{r output-validation}
critical_checks <- tibble(
  check = c(
    "community_metadata_loaded",
    "student_metadata_loaded",
    "community_preference_id_overlap_complete",
    "community_preference_unmatched_ids_absent",
    "cohort_description_table_written",
    "alpha_both_cohorts",
    "alpha_preference_models_written",
    "student_stage2_covariates_valid",
    "student_stage_models_not_identical",
    "permanova_results_written",
    "maaslin_results_written",
    "shotgun_species_scope_valid",
    "main_taxon_figure_written_when_hits_exist",
    "metabolomics_results_written",
    "fb_ratio_harmonized_generated"
  ),
  passed = c(
    nrow(meta_community) > 0,
    nrow(meta_student) > 0,
    community_preference_linkage_audit$n[
      community_preference_linkage_audit$dataset == "IDs shared by meta_dummy and preference file"
    ] == community_preference_linkage_audit$n[
      community_preference_linkage_audit$dataset == "meta_dummy"
    ],
    nrow(community_preference_unmatched_ids) == 0,
    file.exists(file.path(TABLES_DIR, "cohort_description_manuscript_table1.csv")),
    dplyr::n_distinct(alpha_main$cohort) == 2,
    nrow(alpha_model_results) > 0 && file.exists(file.path(TABLES_DIR, "alpha_preference_associations.csv")),
    all(LIFESTYLE_CANDIDATES_STUDENT %in% names(student_stage2_full_design)) &&
      all(vapply(student_stage2_full_design[LIFESTYLE_CANDIDATES_STUDENT], is.factor, logical(1))) &&
      all(vapply(student_stage2_full_design[LIFESTYLE_CANDIDATES_STUDENT], function(x) nlevels(droplevels(x)) > 1, logical(1))),
    nrow(student_stage_comparison) > 0 && all(!student_stage_comparison$effectively_identical),
    nrow(permanova_results) > 0,
    nrow(maaslin_results) > 0,
    isTRUE(species_analysis_scope$scope_valid),
    taxon_association_summary$n_fdr_total == 0 || file.exists(file.path(FIGURES_DIR, "taxa_fdr_significant_main.pdf")),
    nrow(metabolomics_results) > 0,
    !RUN_FB_RATIO || (
      nrow(fb_ratio_diagnostics) == 2 &&
      all(fb_ratio_diagnostics$n_bacillota_rows > 0) &&
      all(fb_ratio_diagnostics$n_bacteroidota_rows > 0) &&
      nrow(fb_preference_results) > 0 &&
      file.exists(file.path(TABLES_DIR, "fb_ratio_preference_associations.csv"))
    )
  )
)

save_table(critical_checks, "analysis_critical_checks.csv")
knitr::kable(critical_checks)

output_manifest <- tibble(
  file = list.files(RUN_DIR, recursive = TRUE, full.names = TRUE),
  relative_path = sub(paste0("^", normalizePath(RUN_DIR), "/?"), "", normalizePath(file, mustWork = FALSE)),
  bytes = file.info(file)$size
) %>%
  dplyr::arrange(relative_path)
save_table(output_manifest, "output_manifest.csv")

ANALYSIS_CHECKS_PASSED <- all(critical_checks$passed)
```

```{r session-info}
sink(file.path(LOG_DIR, "sessionInfo.txt"))
print(sessionInfo())
sink()
sessionInfo()
```

```{r final-check-enforcement}
if (STRICT_CHECKS && !ANALYSIS_CHECKS_PASSED) {
  stop(
    "One or more critical analysis checks failed: ",
    paste(critical_checks$check[!critical_checks$passed], collapse = ", "),
    call. = FALSE
  )
}
message("Completed ", NOTEBOOK_VERSION, ". Outputs: ", RUN_DIR)
```
